---
title: "LapNet QC - étude des données avant DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

<!--
SPDX-FileCopyrightText: 2026 Eugenie Modolo <eugenie.modolo@lyon.unicancer.fr>

SPDX-License-Identifier: AGPL-3.0-or-later
-->

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Lapnet/scripts/Mehlen_LapNet/"
# setwd(path)
knitr::opts_knit$set(root.dir = path)

# files not specific to this project
gft_ref = "/home/eugenie-modolo/Documents/Reference_files/gencode.v49.primary_assembly.annotation.gtf"
mapping_lapnet = "/home/eugenie-modolo/Documents/Reference_files/gene_symbol_mapping_Lapnet.csv"

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")

library("dplyr")
library("viridis")
library("ggplot2")
library("factoextra")
library("reshape2")
library("tidyr")
library("splines")
library(reshape2)
library(estimate)
library(tibble)
library(txdbmaker)
library(conflicted)
library(MCPcounter) # pour aide Mcp-counter: ?MCPcounter.estimate

'%!in%' <- function(x,y)!('%in%'(x,y))

```

# Objectif

Etudier la QC sur les 38 échantillons LapNet (32 échantillons de patients et 6 contrôles), avant recherche de gènes différentiellement exprimés via DESeq2.<br/>
Remarque : comme un seul individu a une progression de type PD, nous allons grouper cette catégorie avec les individus SD, car le PD devrait 'tirer' les SD vers les moins bons pronostics.<br/>
<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./data/colData_LapNet.csv', sep=";")
coldata$Best_response_2 <- case_when(
  coldata$Best_response == "SD" ~ "SD_PD",
  coldata$Best_response == "PD" ~ "SD_PD", 
  coldata$Best_response == "PR" ~ "PR",
  TRUE ~ NA_character_  # for any unexpected values
)
coldata$Type <- case_when(
  startsWith(coldata$SampleID, "Ctrl") == TRUE ~ "control",
  startsWith(coldata$SampleID, "Ctrl") == FALSE ~ "lapnet")

# Adding the Splines columns for tumor size progression
# 0. Normalize tumor size at C0
source("R_functions/normalize_by_C0.R")
coldata <- normalize_by_C0(coldata)

# 1. Fit natural splines for each patient
source("R_functions/calculate_tumor_spline.R")
coldata <- calculate_tumor_spline(coldata, df = 2)

coldata <- subset( coldata, select = - c(Size_lesion_C0, Size_lesion_C4, Size_lesion_C8, Size_lesion_C12, Days_lesion_C0, Days_lesion_C4, Days_lesion_C8, Days_lesion_C12, Best_response, Size_norm_C0))


# scale and center the numeric columns
numeric_columns <- c('Age', 'Height', 'Weight', 'BMI', 'Size_norm_C4', 'Size_norm_C8', 'Size_norm_C12', 'best_lesion_perc', 'OS', 'PFS', 'Surgery', 'QC_duplication_perc', 'QC_alignedM_Salmon', 'QC_uniq_aligned', 'Purity_Estimate', 'Purity_PUREE', 'tumor_spline_slope', 'tumor_spline_curvature')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns 
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)

# countData (gene counts, not transcript's)
counts_genes <- read.table("./data/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
# we only study the genes with expression not always null
counts_genes <- counts_genes[rowSums(counts_genes[, count_columns] != 0) > 0, ]
count_matrix <- as.matrix(counts_genes[, count_columns])

coldata <- data.frame(coldata, row.names = count_columns)

# adding gene length to counts_genes
txdb <- txdbmaker::makeTxDbFromGFF(gft_ref,format="gtf")
# collect the exons per gene id
exons.list.per.gene <- exonsBy(txdb,by="gene")
# then for each gene, reduce all the exons to a set of non overlapping exons, calculate their lengths (widths) and sum then
exonic.gene.sizes <- as.data.frame(sum(width(GenomicRanges::reduce(exons.list.per.gene))))
match_index <- match(rownames(counts_genes), rownames(exonic.gene.sizes))
counts_genes$gene_length <- exonic.gene.sizes[match_index, 1]

# keeping only the protein-coding genes, as we used a protocol capturing only exons
gtf <- import(gft_ref)
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[which(rownames(counts_genes) %in% keep_genes), ]
count_matrix_pcg <- as.matrix(count_matrix_pcg[, count_columns])

# Working with gene symbols
mapping <- read.csv(mapping_lapnet, stringsAsFactors = FALSE)
source("R_functions/apply_gene_symbol_mapping.R")
count_matrix_genesymbol <- apply_gene_symbol_mapping(
  count_matrix,
  mapping_file = mapping,
  df_genelengths = counts_genes
)
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_genesymbol[, count_columns],
  colData = coldata,
  design = ~ 1
)
dds <- DESeq(dds)
norm.count_genesymbol <- as.data.frame(counts(dds, normalized=TRUE))
gene_lengths <- count_matrix_genesymbol$gene_lengths
norm.count_genesymbol_K <- as.matrix(norm.count_genesymbol/gene_lengths)

count_matrix_pcg_genesymbol <- apply_gene_symbol_mapping(
  count_matrix_pcg,
  mapping_file = mapping,
  df_genelengths = counts_genes
)
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_pcg_genesymbol[, count_columns],
  colData = coldata,
  design = ~ 1
)
dds <- DESeq(dds)
norm.count_pcg_genesymbol <- as.data.frame(counts(dds, normalized=TRUE))
gene_lengths <- count_matrix_pcg_genesymbol$gene_lengths
norm.count_pcg_genesymbol_K <- as.matrix(norm.count_pcg_genesymbol/gene_lengths)

genes_of_interest <- c('NTN1', 'NEO1') 
expression_genes <- norm.count_pcg_genesymbol_K[match(genes_of_interest, row.names(norm.count_pcg_genesymbol_K)),]
t_expression_genes <- data.frame(t(expression_genes))
for (gene in genes_of_interest) {
  coldata[[paste0(gene, "_log")]] <- log2(t_expression_genes[[gene]] + 1)
}

pivot_longer(coldata[,c("Type", 'NTN1_log', 'NEO1_log')], cols=c(NTN1_log, NEO1_log)) %>%
  ggplot(aes(x=name,y=value,fill=Type))+
  geom_boxplot() +
  ggtitle("Comparison of expressions between control and LapNet samples")

```

<br/>
<br/>
Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix)` gènes, dont `r nrow(count_matrix_pcg)` protein-coding gènes. Pour la suite, on ne va travailler que sur les protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(coldata) == colnames(count_matrix_pcg)) == length(rownames(coldata))`.<br/>


# Analyses sur le jeu de données

```{r data overview}

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_pcg,
  colData = coldata,
  design = ~ Time
)

dds <- DESeq(dds)
# norm.count_gynet <- as.data.frame(counts(dds, normalized=TRUE))

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Time'): Cooks' distance")

```



## PCAs sur tous les protein-coding gènes : {.tabset}

### axis 1/2

```{r PCA both all ax1-2 v1, fig.width=13, fig.height=10}

# rld <- rlog(dds, blind=TRUE) # too slow for that many samples
rld <- vst(dds, blind=TRUE)

myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
to_be_plotted <- c("SampleID", "Time", "Type", "Age", "Sex", "best_lesion_perc", "tumor_spline_slope", "OS", "Death_status", "PFS", "Prog_status", "Surg_status", "QC_duplication_perc", "QC_alignedM_Salmon", "QC_uniq_aligned", "Best_response_2")
nber_top_genes <- nrow(count_matrix_pcg)

source("R_functions/create_myPCAplots.R")
plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA both all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA both all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA both all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA both all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA both all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA both all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA both all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables associated to PC axis 
```{r PCA both all top variables}

# assess variable contribution to PCs
source("R_functions/assess_variable_contribution.R")
var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


### Top variables contribution with PLS 
```{r PCA both all top variables PLS}

# assess variable importance using PLS
source("R_functions/assess_variable_pls.R")
pls_results <- assess_variable_pls(coldata, rld, ntop = nber_top_genes, ncomp = 2)

# Plot variable importance for both first components
source("R_functions/plot_pls_importance.R")
plot_pls_importance(pls_results)

```

<br/>
<br/>



## PCAs sur les gènes les plus variants : {.tabset}

### choosing optimal number of top variable genes
```{r PCA both variants optimal ntop}

# Calculate variance explained by top N genes, for different gene numbers
gene_numbers <- c(seq(100, 1000, by = 100), seq(1500, 3000, by = 500))
source("R_functions/calculate_variance_by_ngenes.R")
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

On peut choisir 800 gènes, c'est là où la pente de la courbe commence à être moins forte (méthode elbow). 

<br/>
<br/>

### axis 1/2

```{r PCA both variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 800

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA both variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA both variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA both variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA both variants scree plot}

# Calculate PCA of most variant genes
rv <- MatrixGenerics::rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA both variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA both variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA both variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables contribution to PC variance 
```{r PCA both variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

### Top variables contribution with PLS 
```{r PCA both variants top variables PLS}

pls_results <- assess_variable_pls(coldata, rld, ntop = nber_top_genes, ncomp = 2)
plot_pls_importance(pls_results)

```

<br/>
<br/>


## Hierarchical Clustering Heatmaps :  {.tabset}

### Sur tous les protein-coding gènes

```{r heatmaps both all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

# Autre méthode pour plotter les Hierarchical clustering, donne des résultats différents (https://github.com/hamidghaedi/RNA-seq-differential-expression) :
# sampleDists <- dist(t(rld_mat))
# sampleDistMatrix <- as.matrix(sampleDists)
# rownames(sampleDistMatrix) <- colnames(rld_mat)
# colnames(sampleDistMatrix) <- NULL
# colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
# pheatmap(sampleDistMatrix,
#           clustering_distance_rows=sampleDists,
#           clustering_distance_cols=sampleDists,
#           col=colors)

```

<br/>
<br/>

### sur les gènes les plus variants

```{r heatmaps both variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```



# Analyses sur le jeu de données LapNet (sans contrôles) sans les deux outliers (06-006_pre & 07-003_post)

```{r data overview LapNet}

coldata_lapnet <- coldata[which(coldata$Type == "lapnet" & rownames(coldata) != "06-006_pre" & rownames(coldata) != "07-003_post"),]
coldata_lapnet <- subset( coldata_lapnet, select = -c(Type) ) 
count_matrix_lapnet <- count_matrix_pcg[, which(coldata$Type == "lapnet" & rownames(coldata) != "06-006_pre" & rownames(coldata) != "07-003_post")]

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_lapnet,
  colData = coldata_lapnet,
  design = ~ Time
)

dds <- DESeq(dds)

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Time'): Cooks' distance")
# Cook’s distance is a measure of how much a single sample is influencing the fitted coefficients for a gene, and a large value of Cook’s distance is intended to indicate an outlier count

```

## PCAs sur tous les protein-coding gènes : {.tabset}

### axis 1/2

```{r PCA LapNet all ax1-2 v1, fig.width=13, fig.height=10}

rld <- vst(dds, blind=TRUE)

nber_top_genes <- nrow(count_matrix_lapnet)
to_be_plotted <- c("SampleID", "Time", "Size_norm_C4", "Age", "Sex", "best_lesion_perc", "tumor_spline_slope", "OS", "Death_status", "PFS", "Prog_status", "Surg_status", "QC_duplication_perc", "QC_alignedM_Salmon", "QC_uniq_aligned", "Best_response_2")

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA LapNet all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA LapNet all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA LapNet all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA LapNet all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA LapNet all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA LapNet all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA LapNet all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables associated to PC axis 
```{r PCA LapNet all top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


### Top variables contribution with PLS 
```{r PCA LapNet all top variables PLS}

pls_results <- assess_variable_pls(coldata_lapnet, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>



## PCAs sur les gènes les plus variants : {.tabset}

### choosing optimal number of top variable genes
```{r PCA LapNet variants optimal ntop}

# Calculate for different gene numbers
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

On peut choisir 800 gènes, c'est là où la pente de la courbe commence à être moins forte (méthode elbow). 

<br/>
<br/>

### axis 1/2

```{r PCA LapNet variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 800

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA LapNet variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA LapNet variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA LapNet variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_lapnet, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA LapNet variants scree plot}

# Calculate PCA of most variant genes
rv <- MatrixGenerics::rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA LapNet variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA LapNet variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA LapNet variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables contribution to PC variance 
```{r PCA LapNet variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

### Top variables contribution with PLS 
```{r PCA LapNet variants top variables PLS}

pls_results <- assess_variable_pls(coldata_lapnet, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


## Hierarchical Clustering Heatmaps :  {.tabset}

### Sur tous les protein-coding gènes

```{r heatmaps LapNet all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

```

<br/>
<br/>

### sur les gènes les plus variants

```{r heatmaps LapNet variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```





# Analyses sur le jeu de données control

```{r data overview control}

coldata_control <- coldata[which(coldata$Type == "control"),]
coldata_control <- subset( coldata_control, select = - c(Type, Age, Sex, Height, Weight, BMI, Date_traitement, best_lesion_perc, OS, Death_status, PFS, Prog_status, Surgery, Surg_status, Best_response_2, Size_norm_C4, Size_norm_C8, Size_norm_C12, tumor_spline_slope, tumor_spline_curvature, Age_cr, Height_cr, Weight_cr, BMI_cr, Size_norm_C4_cr, Size_norm_C8_cr, Size_norm_C12_cr, best_lesion_perc_cr, OS_cr, PFS_cr, Surgery_cr, tumor_spline_slope_cr, tumor_spline_curvature_cr, Classif_collisson)) # columns that have either only NAs or a unique value
count_matrix_control <- count_matrix_pcg[, which(coldata$Type == "control")]

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_control,
  colData = coldata_control,
  design = ~ Time
)

dds <- DESeq(dds)

# checking the DESeq2 dispersion 
plotDispEsts(dds, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Time'): Cooks' distance")

```

## PCAs sur tous les protein-coding gènes : {.tabset}

### axis 1/2

```{r PCA control all ax1-2 v1, fig.width=13, fig.height=10}

rld <- vst(dds, blind=TRUE)
to_be_plotted <- c("SampleID", "Time", "QC_duplication_perc", "QC_alignedM_Salmon", "QC_uniq_aligned", "Purity_Estimate", "Purity_PUREE", "Classif_moffitt", "Classif_bailey")
nber_top_genes <- nrow(count_matrix_control)

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA control all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA control all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA control all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA control all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA control all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA control all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA control all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables associated to PC axis 
```{r PCA control all top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>


### Top variables contribution with PLS 
```{r PCA control all top variables PLS}

pls_results <- assess_variable_pls(coldata_control, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>



## PCAs sur les gènes les plus variants : {.tabset}

### choosing optimal number of top variable genes
```{r PCA control variants optimal ntop}

# Calculate for different gene numbers
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

Etant donné que la courbe a une forme particulière, on peut choisir 500 gènes (nombre par défaut). 

<br/>
<br/>

### axis 1/2

```{r PCA control variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 500

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA control variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA control variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA control variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(coldata_control, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA control variants scree plot}

# Calculate PCA of most variant genes
rv <- MatrixGenerics::rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA control variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA control variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA control variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables contribution to PC variance 
```{r PCA control variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

### Top variables contribution with PLS 
```{r PCA control variants top variables PLS}

pls_results <- assess_variable_pls(coldata_control, rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


## Hierarchical Clustering Heatmaps :  {.tabset}

### Sur tous les protein-coding gènes

```{r heatmaps control all genes}

rld_mat <- assay(rld)

# rld_cor_p <- cor(rld_mat)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

```

<br/>
<br/>

### sur les gènes les plus variants

```{r heatmaps control variant genes}

rld_mat_500 <- assay(rld)[select, ]

# rld_cor_p <- cor(rld_mat_500)
# pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```






# Conclusions et remarques 

Partie 2 (tout le jeu de données) : deux échantillons ressortent nettement comme outliers : 06-006_pre & 07-003_post. Ce sont ceux qui expliquent le plus la variance sur au moins l'axe 1 des ACP, et ont des scores de corrélation nettement inférieurs à toutes les autres paires d'échantillons (même sur les gènes les plus variants).


Dans la partie 3 (LapNet sans contrôles et sans outliers) : 

* notre variable d'intérêt pre/post (Time) semble séparer sur l'axe 4 des ACP, et elle est plutôt bien placée en PLS (d'autant plus pour les gènes les plus variants) ;

* les variables d'intérêt biomarqueur (OS, PFS, etc.) ne semblent pas séparer sur l'un des 5 premiers axes de l'ACP. En PLS, les mieux placées diffèrent selon si l'on regarde tous les gènes ou uniquement les plus variants. 


Dans la partie 4 (échantillons contrôles uniquement) : étant donné qu'aucune donnée clinique n'est disponible, on ne peut pas utiliser ces échantillons pour trouver les DEGs qui nous intéressent.


Le choix du design sera finalisé avec l'étude des LRT.<br/>



# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
