<!--
SPDX-FileCopyrightText: 2026 Eugenie Modolo <eugenie.modolo@lyon.unicancer.fr>

SPDX-License-Identifier: AGPL-3.0-or-later
-->

---
title: "LapNet DEG - étude des données avant DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Lapnet/"
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")

library("dplyr")
library("ggplot2")
library("factoextra")
library("reshape2")
library("tidyr")
library("splines")

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Etudier les gènes différentiellement exprimés sur les données LapNet.<br/>

Remarque : comme un seul individu a une progression de type PD, nous allons grouper cette catégorie avec les individus SD, car le PD devrait 'tirer' les SD vers les moins bons pronostics.

<br/>
<br/>

Analyse combinée des 23 samples pre et 6 samples post.<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./colData_LapNet.csv', sep=";")
coldata$Best_response_2 <- case_when(
  coldata$Best_response == "SD" ~ "SD_PD",
  coldata$Best_response == "PD" ~ "SD_PD", 
  coldata$Best_response == "PR" ~ "PR",
  TRUE ~ NA_character_  # for any unexpected values
)
# coldata <- tibble(coldata)

# Adding the Splines and AUC columns for tumor size progression
# 0. Normalize tumor size at C0
normalize_by_C0 <- function(coldata,
                            size_cols = c("Size_lesion_C0", "Size_lesion_C4", 
                                        "Size_lesion_C8", "Size_lesion_C12"),
                            time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                        "Days_lesion_C8", "Days_lesion_C12")) {
  
  coldata$Size_norm_C0 <- NA
  coldata$Size_norm_C4 <- NA
  coldata$Size_norm_C8 <- NA
  coldata$Size_norm_C12 <- NA
  
  for (i in 1:nrow(coldata)) {
    factor_norm <- 100 / coldata$Size_lesion_C0[i]
    coldata$Size_norm_C0[i] <- 100
    coldata$Size_norm_C4[i] <- coldata$Size_lesion_C4[i] * factor_norm
    coldata$Size_norm_C8[i] <- coldata$Size_lesion_C8[i] * factor_norm
    coldata$Size_norm_C12[i] <- coldata$Size_lesion_C12[i] * factor_norm
  }
  return(coldata)
}

coldata <- normalize_by_C0(coldata)

# 1. Fit natural splines for each patient
calculate_tumor_spline <- function(coldata, 
                                   size_cols = c("Size_norm_C0", "Size_norm_C4", 
                                                 "Size_norm_C8", "Size_norm_C12"),
                                   time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                                 "Days_lesion_C8", "Days_lesion_C12"),
                                   df = 2) {  # degrees of freedom for spline
  
  coldata$tumor_spline_slope <- NA
  coldata$tumor_spline_curvature <- NA
  
  for (i in 1:nrow(coldata)) {
    sizes <- as.numeric(coldata[i, size_cols])
    time_points <- as.numeric(coldata[i, time_cols])
    valid_idx <- !is.na(sizes)
    
    if (sum(valid_idx) >= 3) {  # Need at least 3 points for spline
      valid_sizes <- sizes[valid_idx]
      valid_times <- time_points[valid_idx]
      
      tryCatch({
        # Fit spline
        spline_fit <- lm(valid_sizes ~ ns(valid_times, df = min(df, sum(valid_idx) - 1)))
        
        # Extract overall trend (coefficient of first spline basis)
        coldata$tumor_spline_slope[i] <- coef(spline_fit)[2]
        
        # If enough df, get curvature (second derivative approximation)
        if (length(coef(spline_fit)) > 2) {
          coldata$tumor_spline_curvature[i] <- coef(spline_fit)[3]
        }
      }, error = function(e) {
        # Skip if spline fails
      })
    } else if (sum(valid_idx) == 2) {
      # Fall back to linear for 2 points
      valid_sizes <- sizes[valid_idx]
      valid_times <- time_points[valid_idx]
      lm_fit <- lm(valid_sizes ~ valid_times)
      coldata$tumor_spline_slope[i] <- coef(lm_fit)[2]
    }
  }
  
  return(coldata)
}

coldata <- calculate_tumor_spline(coldata, df = 2)

# 2. Calculate AUC (total tumor burden over time)
calculate_tumor_auc <- function(coldata, 
                                size_cols = c("Size_norm_C0", "Size_norm_C4", 
                                              "Size_norm_C8", "Size_norm_C12"),
                                time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                              "Days_lesion_C8", "Days_lesion_C12")) {
  
  coldata$tumor_auc <- NA
  coldata$tumor_auc_normalized <- NA
  
  for (i in 1:nrow(coldata)) {
    sizes <- as.numeric(coldata[i, size_cols])
    time_points <- as.numeric(coldata[i, time_cols])
    valid_idx <- !is.na(sizes)
    
    if (sum(valid_idx) >= 2) {
      valid_sizes <- sizes[valid_idx]
      valid_times <- time_points[valid_idx]
      
      # Calculate AUC using trapezoidal rule
      auc <- 0
      for (j in 1:(length(valid_times) - 1)) {
        auc <- auc + (valid_times[j+1] - valid_times[j]) * 
          (valid_sizes[j] + valid_sizes[j+1]) / 2
      }
      
      coldata$tumor_auc[i] <- auc
      
      # Normalize by time duration
      time_duration <- max(valid_times) - min(valid_times)
      coldata$tumor_auc_normalized[i] <- auc / time_duration
    }
  }
  
  return(coldata)
}

coldata <- calculate_tumor_auc(coldata)

coldata <- subset( coldata, select = - c(Days_lesion_C0, Days_lesion_C4, Days_lesion_C8, Days_lesion_C12, Best_response, Size_norm_C0))

# scale and center the numeric columns
numeric_columns <- c('Age', 'Height', 'Weight', 'BMI', 'Size_lesion_C0', 'Size_lesion_C4', 'Size_lesion_C8', 'Size_lesion_C12', 'best_lesion_perc', 'OS', 'PFS', 'Surgery', 'QC_duplication_perc', 'QC_aligned_M', 'QC_uniq_aligned', 'Size_norm_C4', 'Size_norm_C8', 'Size_norm_C12', 'tumor_spline_slope', 'tumor_spline_curvature', 'tumor_auc', 'tumor_auc_normalized')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
#coldata[paste0(numeric_columns, "_cr")] <- lapply(coldata[numeric_columns], scale)
# convert to factors the non-numeric columns
# other_columns <- c("Days_lesion_C0", "Days_lesion_C4", "Days_lesion_C8", "Days_lesion_C12")
factor_columns <- colnames(coldata) %!in% numeric_columns # & colnames(coldata) %!in% other_columns
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)


# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)

# keeping only the protein-coding genes, as we used a protocol capturing only exons
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[match(keep_genes, rownames(counts_genes)), ]

count_matrix <- as.matrix(count_matrix_pcg[, count_columns])
# # Working with gene symbols
# rownames(count_matrix) <- count_matrix_pcg$gene_name  # Gene names become row names
# # duplicated gene symbols: we make the sum of their raw counts (hypothesis: they code for the same protein function)
# count_matrix <- rowsum(count_matrix, rownames(count_matrix))
# # dup_genes <- count_matrix_pcg[duplicated(count_matrix_pcg$gene_name),'gene_name']
# # rownames(count_matrix_pcg)[count_matrix_pcg$gene_name == 'TUBB']

basic_coldata <- data.frame(coldata, row.names = count_columns)
# basic_coldata <- tibble(coldata, row.names = count_columns) # ne prend pas les row.names ...
# checking that the rows of the metadata table and the columns of the expression matrix must have the same name and same order
# print(c('checkpoint, TRUE if data importation is correct: ', sum(rownames(basic_coldata) == colnames(count_matrix)) == length(rownames(basic_coldata))))

```

```{r functions to visualise Splines (to be run before removing the Days_* columns from coldata)}

# Compare spline vs polynomial vs linear fits side by side
plot_tumor_model_comparison <- function(coldata, 
                                        size_cols = c("Size_norm_C0", "Size_norm_C4", 
                                                      "Size_norm_C8", "Size_norm_C12"),
                                        time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                                      "Days_lesion_C8", "Days_lesion_C12"),
                                        sample_id,
                                        df = 2) {
  
  sizes <- as.numeric(coldata[sample_id, size_cols])
  time_points <- as.numeric(coldata[i, time_cols])
  valid_idx <- !is.na(sizes)
  n_valid <- sum(valid_idx)
  
  if (n_valid < 2) {
    stop("Not enough data points for this sample")
  }
  
  valid_sizes <- sizes[valid_idx]
  valid_times <- time_points[valid_idx]
  
  # Observed data
  obs_data <- data.frame(time = valid_times, size = valid_sizes)
  
  # Prediction times for smooth curves
  pred_times <- seq(min(valid_times), max(valid_times), length.out = 100)
  
  # Fit different models
  all_fits <- data.frame()
  
  # 1. Linear
  fit_linear <- lm(valid_sizes ~ valid_times)
  pred_linear <- predict(fit_linear, newdata = data.frame(valid_times = pred_times))
  all_fits <- rbind(all_fits, data.frame(
    time = pred_times,
    size = pred_linear,
    model = "Linear",
    aic = AIC(fit_linear)
  ))
  
  # 2. Polynomial (if enough points)
  if (n_valid >= 3) {
    fit_poly <- lm(valid_sizes ~ poly(valid_times, 2, raw = TRUE))
    pred_poly <- predict(fit_poly, newdata = data.frame(valid_times = pred_times))
    all_fits <- rbind(all_fits, data.frame(
      time = pred_times,
      size = pred_poly,
      model = "Polynomial",
      aic = AIC(fit_poly)
    ))
  }
  
  # 3. Spline (if enough points)
  if (n_valid >= 3) {
    tryCatch({
      fit_spline <- lm(valid_sizes ~ ns(valid_times, df = min(df, n_valid - 1)))
      pred_spline <- predict(fit_spline, newdata = data.frame(valid_times = pred_times))
      all_fits <- rbind(all_fits, data.frame(
        time = pred_times,
        size = pred_spline,
        model = "Spline",
        aic = AIC(fit_spline)
      ))
    }, error = function(e) {
      # Spline failed
    })
  }
  
  # Find best model (lowest AIC) # renvoit une erreur, pas nécessaire de toute façon
  # best_aic <- all_fits %>%
  #   group_by(model) %>%
  #   summarize(aic = first(aic)) %>%
  #   arrange(aic) %>%
  #   slice(1)
  
  # Plot
  library(dplyr)
  p <- ggplot() +
    geom_line(data = all_fits, aes(x = time, y = size, color = model, linetype = model),
              size = 1) +
    geom_point(data = obs_data, aes(x = time, y = size), 
               color = "black", size = 4) +
    geom_point(data = obs_data, aes(x = time, y = size), 
               color = "red", size = 3) +
    scale_color_manual(values = c("Linear" = "blue", 
                                  "Polynomial" = "green", 
                                  "Spline" = "purple")) +
    scale_linetype_manual(values = c("Linear" = "solid", 
                                     "Polynomial" = "dashed", 
                                     "Spline" = "dotted")) +
    labs(title = paste0(sample_id, " - Model Comparison"),
         # subtitle = paste0("Best model (AIC): ", best_aic$model, " (", 
         #                   round(best_aic$aic, 2), ")"),
         x = "Time (days)", 
         y = "Tumor Size (mm)",
         color = "Model",
         linetype = "Model") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  return(p)
}

# # Compare models for specific patients
# # plot_tumor_model_comparison(coldata, sample_id = "01-001_pre", df = 2)

# Compare models for all patients
# plot_list <- list()
# for (id in rownames(coldata)) {
#   plot_list[[id]] <- plot_tumor_model_comparison(coldata, sample_id = id, df = 2)
# }
# # length(plot_list)
# do.call(grid.arrange, c(plot_list[1:9], ncol = 3))
# do.call(grid.arrange, c(plot_list[10:18], ncol = 3))
# do.call(grid.arrange, c(plot_list[19:27], ncol = 3))
# do.call(grid.arrange, c(plot_list[28:29], ncol = 3))

```

Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(basic_coldata) == colnames(count_matrix)) == length(rownames(basic_coldata))`.<br/>

## Vérifications sur le jeu de données global 

```{r data overview}

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

dds_sex <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = basic_coldata,
  design = ~ Sex
)

dds_sex <- DESeq(dds_sex)

# norm.count <- as.data.frame(counts(dds_sex, normalized=TRUE))

# res_sex <- results(dds_sex, alpha=padj.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
# summary(res_sex)

# checking the DESeq2 dispersion 
plotDispEsts(dds_sex, main="DESeq2 dispersion plot")

# looking at the number of outliers
boxplot(log10(assays(dds_sex)[["cooks"]]), range=0, las=2, main="Number of outliers (attribute 'Sex'): Cooks' distance")

```

## PCAs sur tous les protein-coding gènes : {.tabset}

### axis 1/2

```{r PCA all ax1-2 v1, fig.width=13, fig.height=10}

rld <- rlog(dds_sex, blind=TRUE)

myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
to_be_plotted <- c("SampleID", "Time", "Age", "Sex", "best_lesion_perc", "tumor_spline_slope", "tumor_auc_normalized", "OS", "Death_status", "PFS", "Prog_status", "Surg_status", "QC_duplication_perc", "QC_aligned_M", "QC_uniq_aligned", "Best_response_2")
nber_top_genes <- nrow(count_matrix)

create_myPCAplots <- function(df, column, PCx, PCy, nber_top_genes) {
  pcaData <- plotPCA(rld, intgroup=column, pcsToUse=c(PCx,PCy), returnData=TRUE, ntop=nber_top_genes) #, ntop=1000)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  percent_x <- paste0(percentVar[1], "%")
  percent_y <- paste0(percentVar[2], "%")
  x <- paste0('PC', PCx)
  y <- paste0('PC', PCy)
  if (column == 'SampleID') {
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group, label=rownames(df))) +
    geom_point(size=3) +
    ggtitle(column) +
    # theme(plot.title = element_text(hjust=0.95, vjust=0.15, margin = margin(t = -20, b = 10, r = 5), size=11),
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_text_repel(hjust="top", vjust="top") +
    geom_density2d(alpha=.5) +
    coord_fixed() 
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
  else if (column %in% numeric_columns) {
    sc_color <- scale_colour_gradientn(colours = myPalette(100), limits=c(min(df[,which(colnames(df) == column)]), max(df[,which(colnames(df) == column)])))
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group)) +
    geom_point(size=3) +
    ggtitle(column) +
    # theme(plot.title = element_text(hjust=0.95, vjust=0.15, margin = margin(t = -20, b = 10, r = 5), size=11),
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_density2d(alpha=.5) +
    coord_fixed() +
    sc_color
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
  else {
    pca_plot <- ggplot(pcaData, aes(.data[[x]], .data[[y]], color=group)) +
    geom_point(size=3) +
    ggtitle(column) +
    theme(plot.title = element_text(margin = margin(t = 10, b = -20), size=11), legend.position="none", axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), plot.margin = unit(c(1, 1, 1, 1), "mm")) +
    geom_density2d(alpha=.5) +
    coord_fixed()
    return(list("plot" = pca_plot, "PCx" = percent_x, "PCy" = percent_y))
  }
}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA all ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA all ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA all ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA all scree plot}

rld_mat <- assay(rld)
pca <- prcomp(t(rld_mat))
fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA all contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA all top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
# loadings_df$gene_symbol <- counts_genes$gene_name[match(rownames(loadings_df), rownames(counts_genes))]
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
# ggplot(top_loadings_pc1, aes(x = gene_symbol, y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA all biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 1500  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables associated to PC axis 
```{r PCA all top variables}

# Function to assess variable contribution to PCs
assess_variable_contribution <- function(rld, pca, ntop = 500, variables = NULL) {
  
  # If variables not specified, use all from colData
  if (is.null(variables)) { variables <- colnames(colData(rld)) }
  # # Calculate PCA # already done above, so added it to stdin
  # rv <- rowVars(assay(rld))
  # select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  # pca <- prcomp(t(assay(rld)[select, ]))
  # Get PC coordinates
  pca_data <- data.frame(pca$x[, 1:2])
  pca_data <- cbind(pca_data, colData(rld))
  # Calculate variance explained
  percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100
    # For each variable, calculate correlation/association with PCs
  results_list <- list()
  
  for (var in variables) {
    print(var)
    # Skip if variable doesn't exist or has only one level
    if (!var %in% colnames(pca_data)) next
    var_data <- pca_data[[var]]
    # Check if variable is numeric or categorical
    if (is.numeric(var_data) && !is.factor(var_data)) {
      next  # we want only the centered and reduced variables (factors)
      # # For numeric: calculate correlation
      # cor_pc1 <- cor(pca_data$PC1, var_data, use = "complete.obs")
      # cor_pc2 <- cor(pca_data$PC2, var_data, use = "complete.obs")
      # # R-squared as measure of variance explained
      # r2_pc1 <- cor_pc1^2 * 100
      # r2_pc2 <- cor_pc2^2 * 100
    } else if (is.factor(var_data) && endsWith(var, '_cr')) {
      var_data <- as.numeric(as.vector(pca_data[[var]]))
      cor_pc1 <- cor(pca_data$PC1, var_data, use = "complete.obs")
      cor_pc2 <- cor(pca_data$PC2, var_data, use = "complete.obs")
      # R-squared as measure of variance explained
      r2_pc1 <- cor_pc1^2 * 100
      r2_pc2 <- cor_pc2^2 * 100
    }
    else {
      # For categorical: ANOVA / calculate eta-squared
      var_data <- as.factor(var_data)
      # Skip if only one level
      if (length(levels(var_data)) < 2) next
      # ANOVA for PC1
      aov_pc1 <- aov(pca_data$PC1 ~ var_data)
      ss_total_pc1 <- sum((pca_data$PC1 - mean(pca_data$PC1))^2)
      ss_var_pc1 <- sum(aov_pc1$residuals^2)
      r2_pc1 <- (1 - ss_var_pc1/ss_total_pc1) * 100
      # ANOVA for PC2
      aov_pc2 <- aov(pca_data$PC2 ~ var_data)
      ss_total_pc2 <- sum((pca_data$PC2 - mean(pca_data$PC2))^2)
      ss_var_pc2 <- sum(aov_pc2$residuals^2)
      r2_pc2 <- (1 - ss_var_pc2/ss_total_pc2) * 100
    }
    
    results_list[[var]] <- data.frame(
      variable = var,
      PC1_variance_explained = r2_pc1,
      PC2_variance_explained = r2_pc2
    )
  }
  
  # Combine results
  contribution_df <- do.call(rbind, results_list)
  rownames(contribution_df) <- NULL
  # Sort by PC1 contribution
  contribution_df <- contribution_df[order(contribution_df$PC1_variance_explained, 
                                          decreasing = TRUE), ]
  
  return(list(
    contribution = contribution_df,
    percentVar = percentVar,
    pca_data = pca_data
  ))
}

# Use the function
var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# # View results
# print(var_contribution$contribution)

# # Plot the 2 PCs separately
# # Plot PC1 contributions
# p1 <- ggplot(var_contribution$contribution, 
#              aes(x = reorder(variable, PC1_variance_explained), 
#                  y = PC1_variance_explained)) +
#   geom_bar(stat = "identity", fill = "steelblue") +
#   coord_flip() +
#   labs(title = paste0("Variable Contribution to PC1 (", 
#                      round(var_contribution$percentVar[1], 1), "% total variance)"),
#        x = "Variable",
#        y = "Variance Explained (%)") +
#   theme_minimal()
# 
# # Plot PC2 contributions
# p2 <- ggplot(var_contribution$contribution, 
#              aes(x = reorder(variable, PC2_variance_explained), 
#                  y = PC2_variance_explained)) +
#   geom_bar(stat = "identity", fill = "coral") +
#   coord_flip() +
#   labs(title = paste0("Variable Contribution to PC2 (", 
#                      round(var_contribution$percentVar[2], 1), "% total variance)"),
#        x = "Variable",
#        y = "Variance Explained (%)") +
#   theme_minimal()
# 
# # Display plots
# grid.arrange(p1, p2, nrow = 2)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")


# # compute and print the statistical significance test
# test_variable_significance <- function(rld, ntop = nber_top_genes) {
#   # Calculate PCA
#   rv <- rowVars(assay(rld))
#   select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
#   pca <- prcomp(t(assay(rld)[select, ]))
#   pca_data <- data.frame(pca$x[, 1:2])
#   # Add colData columns
#   variables <- colnames(colData(rld))
#   for (var in variables) {
#     if (var %in% colnames(colData(rld))) {
#       pca_data[[var]] <- colData(rld)[[var]]
#     }
#   }
#   results <- list()
#   for (var in variables) {
#     if (!var %in% colnames(pca_data)) next
#     var_data <- pca_data[[var]]
#     # Remove NAs
#     complete_cases <- complete.cases(pca_data$PC1, pca_data$PC2, var_data)
#     if (sum(complete_cases) < 3) next
#     pc1_clean <- pca_data$PC1[complete_cases]
#     pc2_clean <- pca_data$PC2[complete_cases]
#     var_clean <- var_data[complete_cases]
#     if (is.numeric(var_clean) && !is.factor(var_clean)) {
#       # Check for variation
#       if (sd(var_clean, na.rm = TRUE) == 0) next
#       # Correlation test
#       test_pc1 <- tryCatch(cor.test(pc1_clean, var_clean), error = function(e) NULL)
#       test_pc2 <- tryCatch(cor.test(pc2_clean, var_clean), error = function(e) NULL)
#       if (is.null(test_pc1) || is.null(test_pc2)) next
#       results[[var]] <- data.frame(
#         variable = var,
#         type = "numeric",
#         PC1_pvalue = test_pc1$p.value,
#         PC2_pvalue = test_pc2$p.value,
#         PC1_statistic = test_pc1$estimate,
#         PC2_statistic = test_pc2$estimate,
#         stringsAsFactors = FALSE
#       )
#     } else {
#       # ANOVA test
#       var_factor <- as.factor(var_clean)
#       if (length(levels(var_factor)) < 2) next
#       # Perform ANOVA with error handling
#       aov_pc1 <- tryCatch(aov(pc1_clean ~ var_factor), error = function(e) NULL)
#       aov_pc2 <- tryCatch(aov(pc2_clean ~ var_factor), error = function(e) NULL)
#       if (is.null(aov_pc1) || is.null(aov_pc2)) next
#       # Extract p-values safely
#       sum_pc1 <- summary(aov_pc1)[[1]]
#       sum_pc2 <- summary(aov_pc2)[[1]]
#       # Check if p-value exists
#       if (!"Pr(>F)" %in% colnames(sum_pc1) || nrow(sum_pc1) < 1) next
#       if (!"Pr(>F)" %in% colnames(sum_pc2) || nrow(sum_pc2) < 1) next
#       pval_pc1 <- sum_pc1[["Pr(>F)"]][1]
#       pval_pc2 <- sum_pc2[["Pr(>F)"]][1]
#       fstat_pc1 <- sum_pc1[["F value"]][1]
#       fstat_pc2 <- sum_pc2[["F value"]][1]
#       # Skip if NA
#       if (is.na(pval_pc1) || is.na(pval_pc2)) next
#       results[[var]] <- data.frame(
#         variable = var,
#         type = "categorical",
#         PC1_pvalue = pval_pc1,
#         PC2_pvalue = pval_pc2,
#         PC1_statistic = fstat_pc1,
#         PC2_statistic = fstat_pc2,
#         stringsAsFactors = FALSE
#       )
#     }
#   }
#   if (length(results) == 0) {
#     return(data.frame(variable = character(), type = character(), 
#                      PC1_pvalue = numeric(), PC2_pvalue = numeric(),
#                      PC1_statistic = numeric(), PC2_statistic = numeric()))
#   }
#   return(do.call(rbind, results))
# }
# test_variable_significance(rld, ntop=nber_top_genes)

```

<br/>
<br/>


### Top variables contribution with PLS 
```{r PCA all top variables PLS}

# Function to assess variable importance using PLS
assess_variable_pls <- function(rld, ntop = 500, ncomp = 2) {
  
  # Get expression data
  rv <- rowVars(assay(rld))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  X <- t(assay(rld)[select, ])  # Samples x Genes
  # Prepare Y matrix - convert categorical to dummy variables
  Y_list <- list()
  variable_names <- colnames(basic_coldata)
  
  for (var in variable_names) {
    var_data <- basic_coldata[[var]]
    # Skip if all NA
    if (all(is.na(var_data))) next
    # Skip if ID or equivalent # on ne peut pas prendre les classifications, qui sont déterminées directement sur les gènes présents dans le RNAseq (tourne en rond)
    if (var == "SampleID" | var == "Date_traitement" | startsWith(var, "Classif_") | startsWith(var, "Size_")) next
    if (is.numeric(var_data) && !is.factor(var_data)) {
      next  # we want only the centered and reduced variables (factors)
    } else if (is.factor(var_data) && endsWith(var, '_cr')) {
      # Numeric variable - use as is (but centered and scaled)
        Y_list[[var]] <- as.numeric(as.vector(var_data))
    } else {
      # Categorical - convert to dummy variables
      var_factor <- as.factor(var_data)
      if (length(levels(var_factor)) >= 2) {
        # Create dummy variables (one-hot encoding)
        dummy <- model.matrix(~ var_factor - 1)
        colnames(dummy) <- paste0(var, "_", levels(var_factor))
        Y_list[[var]] <- dummy
      }
    }
  }
  
  # Combine into Y matrix
  Y <- do.call(cbind, Y_list)
  
  # Remove rows with NA
  complete_cases <- complete.cases(X, Y)
  X_clean <- X[complete_cases, ]
  Y_clean <- Y[complete_cases, ]
  
  # Run PLS
  pls_result <- pls(X_clean, Y_clean, ncomp = ncomp, mode = "regression")
  
  # Calculate variable importance for each component
  importance_list <- list()
  
  for (comp in 1:ncomp) {
    # Get loadings for Y variables
    y_loadings <- pls_result$loadings$Y[, comp]
    
    # Calculate proportion of variance explained by each original variable
    var_importance <- data.frame(
      variable = character(),
      loading = numeric(),
      importance = numeric(),
      stringsAsFactors = FALSE
    )
    
    current_col <- 1
    for (var in variable_names) {
      if (var %in% names(Y_list)) {
        n_cols <- ncol(as.matrix(Y_list[[var]]))
        var_cols <- current_col:(current_col + n_cols - 1)
        
        # Sum of squared loadings for this variable
        var_loading <- sum(y_loadings[var_cols]^2)
        
        var_importance <- rbind(var_importance, 
                               data.frame(variable = var,
                                        loading = sqrt(var_loading),
                                        importance = var_loading * 100,
                                        stringsAsFactors = FALSE))
        
        current_col <- current_col + n_cols
      }
    }
    
    importance_list[[paste0("Comp", comp)]] <- var_importance
  }
  
  return(list(
    pls_result = pls_result,
    importance = importance_list,
    var_explained = pls_result$prop_expl_var$X,
    Y_matrix = Y_clean,
    X_matrix = X_clean
  ))
}

# Use the function
pls_results <- assess_variable_pls(rld, ntop = nber_top_genes, ncomp = 2)


# Plot variable importance for both first components
plot_pls_importance <- function(pls_results) {
  # Component 1
  imp1 <- pls_results$importance$Comp1
  # imp1 <- imp1[order(imp1$importance, decreasing = TRUE), ]
  imp1$component <- "Component 1"
    # Component 2
  imp2 <- pls_results$importance$Comp2
  # imp2 <- imp2[order(imp2$importance, decreasing = TRUE), ]
  imp2$component <- "Component 2"
  
  imp_combined <- rbind(imp1, imp2)
  
  ggplot(imp_combined, aes(x = reorder(variable, importance), y = importance, fill = component)) +
    geom_bar(stat = "identity", position = "dodge") +
    coord_flip() +
    scale_fill_manual(values = c("Component 1" = "steelblue", 
                                  "Component 2" = "coral")) +
    labs(title = "Variable Importance - PLS Components 1 and 2",
         x = "Variable",
         y = "Importance (%)",
         fill = "Component") +
    theme_minimal()+
    theme(legend.position = "top")
  
}

plot_pls_importance(pls_results)


# # Plot samples in PLS space
# plot_pls_samples <- function(pls_results, rld, color_by = NULL) {
#   
#   # Get sample scores
#   scores <- as.data.frame(pls_results$pls_result$variates$X)
#   colnames(scores) <- paste0("Comp", 1:ncol(scores))
#   
#   # Add metadata
#   if (!is.null(color_by) && color_by %in% colnames(colData(rld))) {
#     scores$color_var <- colData(rld)[[color_by]][complete.cases(
#       pls_results$X_matrix, pls_results$Y_matrix)]
#   }
#   
#   # Variance explained
#   var_exp <- pls_results$pls_result$prop_expl_var$X
#   
#   # Plot
#   p <- ggplot(scores, aes(x = Comp1, y = Comp2))
#   
#   if (!is.null(color_by) && "color_var" %in% colnames(scores)) {
#     p <- p + geom_point(aes(color = color_var), size = 3)
#   } else {
#     p <- p + geom_point(size = 3, color = "blue")
#   }
#   
#   p <- p +
#     labs(title = "PLS Sample Scores",
#          x = paste0("Component 1 (", round(var_exp[1] * 100, 1), "%)"),
#          y = paste0("Component 2 (", round(var_exp[2] * 100, 1), "%)"),
#          color = color_by) +
#     theme_minimal()
#   
#   return(p)
# }
# 
# # Plot samples colored by a variable
# plot_pls_samples(pls_results, rld, color_by = "Time")

```

<br/>
<br/>



## PCAs sur les gènes les plus variants : {.tabset}

### choosing optimal number of top variable genes
```{r PCA variants optimal ntop}

# rld <- rlog(dds_sex, blind=TRUE)

# Function to calculate variance explained by top N genes
calculate_variance_by_ngenes <- function(rld, gene_numbers = seq(100, 2000, by = 100)) {
  
  # Get all gene variances
  rv <- rowVars(assay(rld))
  results <- data.frame(
    n_genes = integer(),
    variance_explained = numeric()
  )
  
  for (n in gene_numbers) {
    # Select top n genes
    select <- order(rv, decreasing = TRUE)[seq_len(min(n, length(rv)))]
    # Run PCA on these genes
    pca <- prcomp(t(assay(rld)[select, ]))
    # Calculate total variance explained by first 2 PCs
    percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100
    total_var_pc1_pc2 <- sum(percentVar[1:2])
    # Or you can use variance explained by all PCs (should be close to 100%)
    # For selection, typically we care about PC1+PC2
    
    results <- rbind(results, data.frame(
      n_genes = n,
      variance_explained = total_var_pc1_pc2
    ))
  }
  return(results)
}

# Calculate for different gene numbers
gene_numbers <- c(seq(100, 1000, by = 100), seq(1500, 3000, by = 500))
variance_results <- calculate_variance_by_ngenes(rld, gene_numbers)

# Plot
ggplot(variance_results, aes(x = n_genes, y = variance_explained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Variance Explained (PC1+PC2) vs Number of Top Variable Genes",
       x = "Number of Top Variable Genes",
       y = "Variance Explained by PC1+PC2 (%)") +
  theme_minimal() +
  geom_vline(xintercept = 500, linetype = "dashed", color = "gray") +
  annotate("text", x = 500, y = min(variance_results$variance_explained), 
           label = "Default: 500", vjust = -1)

# Find the "elbow" point
#print(variance_results)

```

On peut choisir 600 gènes, c'est là où la pente de la courbe commence à être moins forte (méthode elbow). 

<br/>
<br/>

### axis 1/2

```{r PCA variants ax1-2 v1, fig.width=13, fig.height=10}

nber_top_genes <- 600

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 2, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/3
```{r PCA variants ax1-3 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 3, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/4
```{r PCA variants ax1-4 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 4, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### axis 1/5
```{r PCA variants ax1-5 v1, fig.width=13, fig.height=10}

plot_list <- list()
for (column in to_be_plotted) {
  newlist <- create_myPCAplots(basic_coldata, column, 1, 5, nber_top_genes)
  plot_list[[column]] <- newlist$plot
  perc_expliq_x <- newlist$PCx
  perc_expliq_y <- newlist$PCy
}
grid.arrange(grobs = plot_list, ncol = 4, nrow = 4)

```

% de variance expliquée : axe des X : `r perc_expliq_x`, axe des Y : `r perc_expliq_y`<br/>
Légende pour les variables quantitatives : bleu foncé = valeur min, rouge foncé = valeur max, jaune = milieu

<br/>
<br/>

### scree plot
```{r PCA variants scree plot}

# Calculate PCA of most variant genes
rv <- rowVars(assay(rld))
select <- order(rv, decreasing = TRUE)[seq_len(min(nber_top_genes, length(rv)))]
pca <- prcomp(t(assay(rld)[select, ]))

fviz_eig(pca)

```

<br/>
<br/>

### contributions of individual samples to PCs 1-6
```{r PCA variants contribution of samples}

plot_list <- list()
for (axe in 1:6) {
  plot_list[[axe]] <- fviz_contrib(pca, choice = "ind", axes = axe, ggtheme = theme_minimal() + theme(axis.text.x=element_text(size=8, angle=45)) )
}
grid.arrange(grobs = plot_list, ncol=2, nrow=3)

```

<br/>
<br/>

### top genes contributing to PC variance
```{r PCA variants top genes for PC1}

loadings <- pca$rotation
loadings_df <- data.frame(
  gene = rownames(loadings),
  PC1 = loadings[, "PC1"],
  PC2 = loadings[, "PC2"]
)
top_n <- 30

# Plot PC1 loadings
top_loadings_pc1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc1, aes(x = reorder(gene, abs(PC1)), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC1",
       x = "Gene",
       y = "PC1 Loading") +
  theme_minimal()

# Plot PC2 loadings
top_loadings_pc2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE)[1:top_n], ]
ggplot(top_loadings_pc2, aes(x = reorder(gene, abs(PC2)), y = PC2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 30 Genes Contributing to PC2",
       x = "Gene",
       y = "PC2 Loading") +
  theme_minimal()

```

<br/>
<br/>

### Biplots 
```{r PCA variants biplots}

# Overall biplot
fviz_pca_biplot(pca)

# Get sample coordinates
pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample = colnames(rld)
)

# Calculate variance explained by each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Get top gene loadings (scale for visualization)
top_genes <- 30
loading_scale <- 400  # Adjust this to scale arrows appropriately

top_loading_idx <- order(rowSums(abs(loadings[, 1:2])), decreasing = TRUE)[1:top_genes]
gene_loadings <- data.frame(
  PC1 = loadings[top_loading_idx, "PC1"] * loading_scale,
  PC2 = loadings[top_loading_idx, "PC2"] * loading_scale,
  gene = rownames(loadings)[top_loading_idx]
)
gene_loadings$gene_symbol <- counts_genes$gene_name[match(rownames(gene_loadings), rownames(counts_genes))]

# Create biplot
ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2), size = 3, color = "blue") +
  geom_segment(data = gene_loadings, 
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", alpha = 0.7) +
  geom_text(data = gene_loadings, 
            aes(x = PC1, y = PC2, label = gene_symbol),
            size = 3, vjust = -0.5, hjust = 0.5) +
  labs(title = "PCA Biplot - Samples and Top Contributing Genes",
       x = paste0("PC1 (", round(percentVar[1], 1), "%)"),
       y = paste0("PC2 (", round(percentVar[2], 1), "%)")) +
  theme_minimal()

```

<br/>
<br/>

### Top variables contribution to PC variance 
```{r PCA variants top variables}

var_contribution <- assess_variable_contribution(rld, pca, ntop = nber_top_genes)

# Plot the 2 PCs combined
contribution_long <- melt(var_contribution$contribution,
                         id.vars = "variable",
                         variable.name = "PC",
                         value.name = "variance_explained")

contribution_long$PC <- gsub("PC1_variance_explained", "PC1", contribution_long$PC)
contribution_long$PC <- gsub("PC2_variance_explained", "PC2", contribution_long$PC)
# Plot both PCs together
ggplot(contribution_long, aes(x = reorder(variable, variance_explained),
                              y = variance_explained,
                              fill = PC)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "coral")) +
  labs(title = "Variable Contributions to PC1 and PC2",
       x = "Variable",
       y = "Variance Explained (%)") +
  theme_minimal() +
  theme(legend.position = "top")

```

<br/>
<br/>

### Top variables contribution with PLS 
```{r PCA variants top variables PLS}

pls_results <- assess_variable_pls(rld, ntop = nber_top_genes, ncomp = 2)

plot_pls_importance(pls_results)

```

<br/>
<br/>


## Hierarchical Clustering Heatmaps :  {.tabset}

### Sur tous les protein-coding gènes

```{r heatmaps all}

rld_mat <- assay(rld)

rld_cor_p <- cor(rld_mat)
pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat[,1:4])

```

<br/>
<br/>

### sur les gènes les plus variants

```{r heatmaps variants}

rld_mat_500 <- assay(rld)[select, ]

rld_cor_p <- cor(rld_mat_500)
pheatmap(rld_cor_p, main="pairwise Pearson correlations")

rld_cor_s <- cor(rld_mat_500, method="spearman")
pheatmap(rld_cor_s, main="pairwise Spearman correlations")

ggpairs(rld_mat_500[,1:4])

```


## Utilisation du LRT pour choisir le design

```{r LRT}

coldata_wo_rep <- subset( basic_coldata, select = -c(Age, Height, Weight, BMI, Size_lesion_C0, Size_lesion_C4, Size_lesion_C8, Size_lesion_C12, best_lesion_perc, OS, PFS, Surgery, QC_duplication_perc, QC_aligned_M, QC_uniq_aligned, Size_norm_C4, Size_norm_C8, Size_norm_C12, tumor_spline_slope, tumor_auc, tumor_auc_normalized, Classif_collisson, Classif_moffitt, Classif_bailey) )

for (col in colnames(coldata_wo_rep)) {
    if (endsWith(col, '_cr')) {
        coldata_wo_rep[[col]] <- as.numeric(as.vector(coldata_wo_rep[[col]]))
    }
}

# Function to check if a design matrix is full rank
check_design_rank <- function(count_matrix, coldata, formula) {
  
  tryCatch({
    # Create model matrix
    mm <- model.matrix(formula, data = coldata)
    # Check rank
    is_full_rank <- qr(mm)$rank == ncol(mm)
    
    return(list(
      is_valid = is_full_rank,
      rank = qr(mm)$rank,
      ncol = ncol(mm)
    ))
  }, error = function(e) {
    return(list(
      is_valid = FALSE,
      error = e$message
    ))
  })
}

# Main function that creates dds objects as needed
test_variable_influence_lrt <- function(count_matrix,
                                        coldata,
                                        variables = NULL,
                                        padj_cutoff = 0.05,
                                        lfc_cutoff = 0.58,
                                        max_combination_size = 3) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    # Try to identify likely variables (exclude sample names, IDs, etc.)
    variables <- colnames(coldata)
    # cat("Available variables in coldata:", paste(variables, collapse = ", "), "\n")
    # cat("Testing all variables\n")
  } else {
    # Check that all specified variables exist
    missing_vars <- setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  
  # cat("\nVariables to test:", paste(variables, collapse = ", "), "\n")
  # cat("padj cutoff:", padj_cutoff, "\n")
  # cat("LFC cutoff:", lfc_cutoff, "\n\n")
  
  # Storage for results
  results_list <- list()
  skipped_tests <- list()
  
  # 1. Test each variable alone (reduced = ~1)
  # cat("=== Testing individual variables ===\n")
  for (var in variables) {
    # cat("Testing:", var, "vs ~1\n")
    
    # Check design validity
    formula_full <- as.formula(paste("~", var))
    rank_check <- check_design_rank(count_matrix, coldata, formula_full)
    
    if (!rank_check$is_valid) {
      # cat("  SKIPPED: Design matrix not full rank\n")
      skipped_tests[[paste(var, "vs", "~1")]] <- "Not full rank"
      next
    }
    
    tryCatch({
      # Create dds for this test
      dds <- DESeqDataSetFromMatrix(
        countData = count_matrix,
        colData = coldata,
        design = formula_full
      )
      
      # Run LRT
      dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ 1, quiet = TRUE)
      res <- results(dds_lrt)
      
      # Count DEGs
      n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
      n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                       na.rm = TRUE)
      
      results_list[[var]] <- data.frame(
        full_model = var,
        reduced_model = "~1",
        test_type = "single_variable",
        tested_variable = var,
        n_variables = 1,
        n_deg_padj_only = n_deg_padj,
        n_deg_padj_and_lfc = n_deg_both,
        stringsAsFactors = FALSE
      )
      
      # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
      
    }, error = function(e) {
      cat("  ERROR:", e$message, "\n")
      skipped_tests[[paste(var, "vs", "~1")]] <- e$message
    })
  }
  
  # 2. Test pairs of variables (full = A + B, reduced = A or B)
  if (length(variables) >= 2 && max_combination_size >= 2) {
    # cat("\n=== Testing pairs of variables ===\n")
    
    for (i in 1:(length(variables) - 1)) {
      for (j in (i + 1):length(variables)) {
        var1 <- variables[i]
        var2 <- variables[j]
        
        # Test A + B vs A (to see effect of B)
        # cat("Testing:", var1, "+", var2, "vs", var1, "\n")
        
        formula_full <- as.formula(paste("~", var1, "+", var2))
        formula_reduced <- as.formula(paste("~", var1))
        
        # Check design validity
        rank_check_full <- check_design_rank(count_matrix, coldata, formula_full)
        rank_check_reduced <- check_design_rank(count_matrix, coldata, formula_reduced)
        
        if (!rank_check_full$is_valid || !rank_check_reduced$is_valid) {
          # cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                           na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var1, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var1,
            test_type = "added_variable",
            tested_variable = var2,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
          
        }, error = function(e) {
          # cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- e$message
        })
        
        # Test A + B vs B (to see effect of A)
        # cat("Testing:", var1, "+", var2, "vs", var2, "\n")
        
        formula_reduced2 <- as.formula(paste("~", var2))
        rank_check_reduced2 <- check_design_rank(count_matrix, coldata, formula_reduced2)
        
        if (!rank_check_full$is_valid || !rank_check_reduced2$is_valid) {
          # cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced2, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                           na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var2, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var2,
            test_type = "added_variable",
            tested_variable = var1,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
          
        }, error = function(e) {
          cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- e$message
        })
      }
    }
  }
  
  # 3. Test triplets (full = A + B + C, reduced = A + B)
  if (length(variables) >= 3 && max_combination_size >= 3) {
    # cat("\n=== Testing triplets of variables ===\n")
    
    for (i in 1:(length(variables) - 2)) {
      for (j in (i + 1):(length(variables) - 1)) {
        for (k in (j + 1):length(variables)) {
          var1 <- variables[i]
          var2 <- variables[j]
          var3 <- variables[k]
          
          # cat("Testing:", var1, "+", var2, "+", var3, "vs", var1, "+", var2, "\n")
          
          formula_full <- as.formula(paste("~", var1, "+", var2, "+", var3))
          formula_reduced <- as.formula(paste("~", var1, "+", var2))
          
          # Check design validity
          rank_check_full <- check_design_rank(count_matrix, coldata, formula_full)
          rank_check_reduced <- check_design_rank(count_matrix, coldata, formula_reduced)
          
          if (!rank_check_full$is_valid || !rank_check_reduced$is_valid) {
            # cat("  SKIPPED: Design matrix not full rank\n")
            skipped_tests[[paste(var1, var2, var3, "vs", var1, var2)]] <- "Not full rank"
            next
          }
          
          tryCatch({
            dds <- DESeqDataSetFromMatrix(
              countData = count_matrix,
              colData = coldata,
              design = formula_full
            )
            
            dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced, quiet = TRUE)
            res <- results(dds_lrt)
            
            n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
            n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                             na.rm = TRUE)
            
            results_list[[paste(var1, var2, var3, "vs", var1, var2, sep = "_")]] <- data.frame(
              full_model = paste(var1, "+", var2, "+", var3),
              reduced_model = paste(var1, "+", var2),
              test_type = "added_to_pair",
              tested_variable = var3,
              n_variables = 3,
              n_deg_padj_only = n_deg_padj,
              n_deg_padj_and_lfc = n_deg_both,
              stringsAsFactors = FALSE
            )
            
            # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
            
          }, error = function(e) {
            cat("  ERROR:", e$message, "\n")
            skipped_tests[[paste(var1, var2, var3, "vs", var1, var2)]] <- e$message
          })
        }
      }
    }
  }
  
  # Combine all results
  if (length(results_list) == 0) {
    cat("\nWARNING: No tests were successful!\n")
    results_df <- data.frame()
  } else {
    results_df <- do.call(rbind, results_list)
    rownames(results_df) <- NULL
  }
  
  # Summary
  cat("\n=== SUMMARY ===\n")
  cat("Successful tests:", length(results_list), "\n")
  cat("Skipped/failed tests:", length(skipped_tests), "\n")
  
  if (length(skipped_tests) > 0) {
    cat("\nSkipped tests:\n")
    for (test_name in names(skipped_tests)) {
      cat("  ", test_name, ":", skipped_tests[[test_name]], "\n")
    }
  }
  
  return(list(
    results = results_df,
    skipped = skipped_tests
  ))
}

# Run the analysis 
lrt_results <- test_variable_influence_lrt(
  count_matrix = count_matrix,
  coldata = coldata_wo_rep,
  variables = c("Time", "Sex", "Death_status", "Prog_status", "Surg_status", "Best_response_2", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "best_lesion_perc_cr", "OS_cr", "PFS_cr", "QC_duplication_perc_cr", "QC_aligned_M_cr", "QC_uniq_aligned_cr", "tumor_spline_slope_cr", "tumor_auc_cr", "tumor_auc_normalized_cr"),  # or NULL for all # je retire les colonnes d'identifiants uniques (SampleID et Date_traitement), celles avec des NA (DESeq2 fait une erreur sinon) et celles qui sont des répétitions (pour raccourcir le temps de calcul)
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58,
  max_combination_size = 2  # Test singles and pairs only
)


# draw the results table
as.matrix(lrt_results$results) %>%
DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
DT::formatStyle(columns = colnames(lrt_results$results),fontSize = '80%')

# if tests were skipped, draw them as well
if (length(lrt_results$skipped) > 0) {
  skipped_df <- data.frame(
    test = names(lrt_results$skipped),
    reason = unlist(lrt_results$skipped)
  )
  as.matrix(skipped_df) %>%
  DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
  DT::formatStyle(columns = colnames(skipped_df),fontSize = '80%')
}


# Create symmetric heatmap showing maximum effect in either direction
library(viridis)
plot_symmetric_deg_heatmap <- function(lrt_results, criterion = "padj_and_lfc") {
  
  results_df <- lrt_results$results
  pairwise <- results_df[results_df$test_type == "added_variable", ]
  
  if (nrow(pairwise) == 0) {
    stop("No pairwise tests found in results")
  }
  
  # Choose DEG count
  if (criterion == "padj_only") {
    pairwise$n_deg <- pairwise$n_deg_padj_only
    legend_title <- "DEGs (padj only)"
  } else {
    pairwise$n_deg <- pairwise$n_deg_padj_and_lfc
    legend_title <- "DEGs (padj + LFC)"
  }
  
  # Get all unique variables
  all_vars <- unique(c(pairwise$reduced_model, pairwise$tested_variable))
  
  # Create symmetric matrix
  deg_matrix <- matrix(NA, nrow = length(all_vars), ncol = length(all_vars))
  rownames(deg_matrix) <- all_vars
  colnames(deg_matrix) <- all_vars
  
  # Fill matrix - for each pair, take the number of DEGs
  for (i in 1:nrow(pairwise)) {
    row_var <- pairwise$tested_variable[i]
    col_var <- pairwise$reduced_model[i]
    n_degs <- pairwise$n_deg[i]
    deg_matrix[row_var, col_var] <- n_degs
    
    # # Fill both directions with the maximum DEGs from either direction
    # current_val <- deg_matrix[row_var, col_var]
    # if (is.na(current_val) || n_degs > current_val) {
    #   deg_matrix[row_var, col_var] <- n_degs
    #   deg_matrix[col_var, row_var] <- n_degs  # Symmetric
    # }
  }
  
  # Diagonal = single variable tests
  single_var <- results_df[results_df$test_type == "single_variable", ]
  if (nrow(single_var) > 0) {
    for (i in 1:nrow(single_var)) {
      var <- single_var$full_model[i]
      if (var %in% all_vars) {
        if (criterion == "padj_only") {
          deg_matrix[var, var] <- single_var$n_deg_padj_only[i]
        } else {
          deg_matrix[var, var] <- single_var$n_deg_padj_and_lfc[i]
        }
      }
    }
  }
  
  # Convert to long format
  heatmap_long <- melt(deg_matrix, na.rm = FALSE)
  colnames(heatmap_long) <- c("variable1", "variable2", "n_deg")
  
  # Plot
  p <- ggplot(heatmap_long, aes(x = variable2, y = variable1, fill = n_deg)) +
    geom_tile(color = "white", size = 0.5) +
    geom_text(aes(label = ifelse(is.na(n_deg), "", as.character(n_deg))), 
              color = "white", size = 3.5, fontface = "bold") +
    scale_fill_viridis(option = "plasma", na.value = "grey90",
                      name = legend_title,
                      limits = c(0, max(heatmap_long$n_deg, na.rm = TRUE))) +
    labs(title = "Pairwise Variable Effects on Differential Expression",
         subtitle = "Diagonal: single variable | Off-diagonal: DEGs for the LRT tested on the row",
         x = "Variable",
         y = "Variable") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, size = 9),
          legend.position = "right",
          panel.grid = element_blank())
  
  return(p)
}

# Use it
plot_symmetric_deg_heatmap(lrt_results, criterion = "padj_only") 
plot_symmetric_deg_heatmap(lrt_results, criterion = "padj_and_lfc")

```

```{r function LRT with design0}

test_variable_influence_lrt_design0 <- function(count_matrix,
                                        coldata,
                                        variables = NULL,
                                        padj_cutoff = 0.05,
                                        lfc_cutoff = 0.58,
                                        design0 = 'Time',
                                        max_combination_size = 2) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    # Try to identify likely variables (exclude sample names, IDs, etc.)
    variables <- colnames(coldata)
    # cat("Available variables in coldata:", paste(variables, collapse = ", "), "\n")
    # cat("Testing all variables\n")
  } else {
    # Check that all specified variables exist
    missing_vars <- setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  
  # cat("\nVariables to test:", paste(variables, collapse = ", "), "\n")
  # cat("padj cutoff:", padj_cutoff, "\n")
  # cat("LFC cutoff:", lfc_cutoff, "\n\n")
  
  # Storage for results
  results_list <- list()
  skipped_tests <- list()
  
  # 1. Test each variable alone (reduced = ~design0)
  cat("=== Testing individual variables ===\n")
  for (var in variables) {
    cat("Testing:", var, "vs design0\n")
    
    # Check design validity
    formula_full <- as.formula(paste("~ ", var, " + ", design0))
    rank_check <- check_design_rank(count_matrix, coldata, formula_full)
    
    if (!rank_check$is_valid) {
      cat("  SKIPPED: Design matrix not full rank\n")
      skipped_tests[[paste(var, "vs", "design0")]] <- "Not full rank"
      next
    }
    
    tryCatch({
      # Create dds for this test
      dds <- DESeqDataSetFromMatrix(
        countData = count_matrix,
        colData = coldata,
        design = formula_full
      )
      
      # Run LRT
      dds_lrt <- DESeq(dds, test = "LRT", reduced = as.formula(paste("~ ", design0)), quiet = TRUE)
      res <- results(dds_lrt)

      # Count DEGs
      n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
      n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                        na.rm = TRUE)
      
      results_list[[var]] <- data.frame(
        full_model = var,
        reduced_model = "~1",
        test_type = "single_variable",
        tested_variable = var,
        n_variables = 1,
        n_deg_padj_only = n_deg_padj,
        n_deg_padj_and_lfc = n_deg_both,
        stringsAsFactors = FALSE
      )
      
      cat("  DEGs (padj):", n_deg_padj, "\n")
      
    }, error = function(e) {
      cat("  ERROR:", e$message, "\n")
      skipped_tests[[paste(var, "vs", "design0")]] <- e$message
    })
  }
  
  # 2. Test pairs of variables (full = design0 + A + B, reduced = design0 + A or B)
  if (length(variables) >= 2 && max_combination_size >= 2) {
    cat("\n=== Testing pairs of variables ===\n")
    
    for (i in 1:(length(variables) - 1)) {
      for (j in (i + 1):length(variables)) {
        var1 <- variables[i]
        var2 <- variables[j]
        
        # Test A + B vs A (to see effect of B)
        cat("Testing:", var1, "+", var2, "vs", var1, "\n")
        
        formula_full <- as.formula(paste("~", var1, "+", var2, " + ", design0))
        formula_reduced <- as.formula(paste("~", var1, " + ", design0))
        
        # Check design validity
        rank_check_full <- check_design_rank(count_matrix, coldata, formula_full)
        rank_check_reduced <- check_design_rank(count_matrix, coldata, formula_reduced)
        
        if (!rank_check_full$is_valid || !rank_check_reduced$is_valid) {
          cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                            na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var1, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var1,
            test_type = "added_variable",
            tested_variable = var2,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          cat("  DEGs (padj):", n_deg_padj, "\n")
          
        }, error = function(e) {
          cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- e$message
        })
        
        # Test A + B vs B (to see effect of A)
        cat("Testing:", var1, "+", var2, "vs", var2, "\n")
        
        formula_reduced2 <- as.formula(paste("~", var2, " + ", design0))
        rank_check_reduced2 <- check_design_rank(count_matrix, coldata, formula_reduced2)
        
        if (!rank_check_full$is_valid || !rank_check_reduced2$is_valid) {
          cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced2, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                            na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var2, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var2,
            test_type = "added_variable",
            tested_variable = var1,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          cat("  DEGs (padj):", n_deg_padj, "\n")
          
        }, error = function(e) {
          cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- e$message
        })
      }
    }
  }
  
  # Combine all results
  if (length(results_list) == 0) {
    cat("\nWARNING: No tests were successful!\n")
    results_df <- data.frame()
  } else {
    results_df <- do.call(rbind, results_list)
    rownames(results_df) <- NULL
  }
  
  # Summary
  cat("\n=== SUMMARY ===\n")
  cat("Successful tests:", length(results_list), "\n")
  cat("Skipped/failed tests:", length(skipped_tests), "\n")
  
  if (length(skipped_tests) > 0) {
    cat("\nSkipped tests:\n")
    for (test_name in names(skipped_tests)) {
      cat("  ", test_name, ":", skipped_tests[[test_name]], "\n")
    }
  }
  
  return(list(
    results = results_df,
    skipped = skipped_tests
  ))
}

# # Run the analysis 
# lrt_results <- test_variable_influence_lrt_design0(
#   count_matrix = count_matrix,
#   coldata = coldata_wo_rep,
#   variables = c("SampleID", "Sex", "Death_status", "Prog_status", "Surg_status", "Best_response_2", "Height_cr", "Weight_cr", "BMI_cr", "best_lesion_perc_cr", "OS_cr", "PFS_cr", "QC_duplication_perc_cr", "QC_uniq_aligned_cr", "tumor_spline_slope_cr", "tumor_auc_cr", "tumor_auc_normalized_cr"),
#   padj_cutoff = 0.05,
#   lfc_cutoff = 0.58,
#   design0 = Age_cr + QC_aligned_M_cr + Time,
#   max_combination_size = 2  # Test singles and pairs only
# )
# 
# plot_symmetric_deg_heatmap(lrt_results, criterion = "padj_only")

```


# Conclusion et remarques

Choix du design, en mettant plus de poids aux LRT (pas impacté par le nombre de variables testées), puis aux PLS pour tous les gènes, et enfin aux PLS des 600 gènes plus variants :<br/>
- variables à inclure pour retirer les effets non recherchés (on en choisit une par catégorie pour qu'elles ne risquent pas d'être orthogonales) : <br/>
  - pour les données cliniques : Age_cr est le mieux placé dans 2 des trois (LRT, PLS_600). Il pourrait être intéressant de tester en plus (si non-orthogonal) : (i) Sex (bien classé en LRT et PLS_all), (ii) Height_cr (bien classé en LRT, mais pas en PLS et pas de logique biologique ...), (iii) les 4 variables de survie citées ci-dessous <br/> 
  - pour les données de QC : QC_aligned_M_cr est le mieux placé dans 2 des trois (LRT, PLS_all) <br/> 
  - pour l'étude des pré : il faudra peut-être ajouter Time dans les variables avant celle d'intérêt, car il est très haut dans LRT mais pas les PLS.<br/>
<br/>
- variables explicatives :<br/>
  - pour la recherche de biomarqueurs de survie (étude uniquement des pre) : il y a le choix entre : (i) best_lesion_perc_cr (très bon en valeur unique dans le LRT et dans PLS_600), (ii) Best_response_2 (bien classé dans les 3), (iii) tumor_spline_slope_cr (bon en LRT mais pas super en PLS), (iv) Prog_status (pas mal classé dans les 3).<br/>
  - pour la différence pre/post : Time ressort très nettement en LRT (c'était la seule variable pour cette condition de toutes façons).<br/>

Remarques sur les samples outliers :<br/>
- pour les heatmap je regarde les Spearman car les pairplot ne montrent pas de corrélation linéaire.<br/>
- 06-008 et 06-009 ressortent comme outlier dans plusieurs figures (heatmap, contribution aux PC2,3,4,5) (moins marqué pour 06-009) pour les 600 gènes plus variants. A surveiller ...<br/>
- 06-006 ne ressort pas comme outlier, malgré sa très mauvaise QC : on doit pouvoir le garder. Mais il est tout de même particulier dans le heatmap (proche de tout le monde) : à surveiller.<br/>
- il n'y a pas vraiment de groupe qui ressorte du heatmap pour tous les gènes, mais il y a deux groupes (avec des posts dans les 2) avec les 600 gènes plus variants.


# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
