<!--
SPDX-FileCopyrightText: 2026 Eugenie Modolo <eugenie.modolo@lyon.unicancer.fr>

SPDX-License-Identifier: AGPL-3.0-or-later
-->

---
title: "LapNet DEG - analyse des DEGs obtenus"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Lapnet/"
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}

# /!\ aller voir : https://ngs101.com/how-to-analyze-rnaseq-data-for-absolute-beginners-part-5-from-degs-to-pathways-best-practices/

# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("ggrepel")
library("apeglm")
library("DEGreport")
library(fgsea)
library(clusterProfiler)

library("dplyr")
library("ggplot2")
library('gplots')
library("gridExtra")
library("factoextra")
library(msigdbr)
library(tibble)
library("ComplexHeatmap")
library(circlize)
# library(genekitr)

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Etudier les gènes différentiellement exprimés sur les données LapNet.<br/> 
Nous retirons de l'étude le sample 06-006, qui a seulement 96% de gènes avec une expression non-nulle.<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./colData_LapNet.csv', sep=";")
coldata$Best_response_2 <- case_when(
  coldata$Best_response == "SD" ~ "SD_PD",
  coldata$Best_response == "PD" ~ "SD_PD", 
  coldata$Best_response == "PR" ~ "PR",
  TRUE ~ NA_character_  # for any unexpected values
)
coldata <- subset( coldata, select = -Best_response )

# scale and center the numeric columns
numeric_columns <- c('Age', 'Height', 'Weight', 'BMI', 'Size_lesion_C0', 'Size_lesion_C4', 'Size_lesion_C8', 'Size_lesion_C12', 'best_lesion_perc', 'OS', 'PFS', 'Surgery', 'QC_duplication_perc', 'QC_aligned_M', 'QC_uniq_aligned')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
coldata <- subset( coldata, select = - c(Age, Height, Weight, BMI, Size_lesion_C0, Size_lesion_C4, Size_lesion_C8, Size_lesion_C12, best_lesion_perc, OS, PFS, Surgery, QC_duplication_perc, QC_aligned_M, QC_uniq_aligned))
#coldata[paste0(numeric_columns, "_cr")] <- lapply(coldata[numeric_columns], scale)
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)

# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)

# keeping only the protein-coding genes, as we used a protocol capturing only exons
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.chr_patch_hapl_scaff.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[match(keep_genes, rownames(counts_genes)), ]

count_matrix <- as.matrix(count_matrix_pcg[, count_columns])
# rownames(count_matrix) <- count_matrix_pcg$gene_name  # Gene names become row names
# duplicated gene symbols: we make the sum of their raw counts (hypothesis: they code for the same protein function)
# count_matrix <- rowsum(count_matrix, rownames(count_matrix))
# dup_genes <- count_matrix_pcg[duplicated(count_matrix_pcg$gene_name),'gene_name']
# rownames(count_matrix_pcg)[count_matrix_pcg$gene_name == 'TUBB']

basic_coldata <- data.frame(coldata, row.names = count_columns)
for (col in colnames(basic_coldata)) {
  var_data <- basic_coldata[[col]]
  if (is.factor(var_data) && endsWith(col, '_cr')) {
    basic_coldata[col] <- as.numeric(as.vector(var_data))
  }
}
# checking that the rows of the metadata table and the columns of the expression matrix must have the same name and same order
# print(c('checkpoint, TRUE if data importation is correct: ', sum(rownames(basic_coldata) == colnames(count_matrix)) == length(rownames(basic_coldata))))

# # Removing columns with NAs for DESeq2
# coldata_wo_NA <- subset( basic_coldata, select = -c(Size_lesion_C8, Size_lesion_C12, Surgery) )

# draw basic_coldata table
as.matrix(basic_coldata) %>%
DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
DT::formatStyle(columns = colnames(basic_coldata),fontSize = '80%')


# removing 06-006 
coldata_28 <- basic_coldata[basic_coldata$SampleID != "06-006",]
count_matrix_28 <- count_matrix[,basic_coldata$SampleID != "06-006"]

coldata_22 <- coldata_28[coldata_28$Time == "pre",]
count_matrix_22 <- count_matrix_28[, coldata_28$Time == "pre"]

coldata_8 <- coldata_28[which(coldata_28$SampleID %in% c("01-007", "02-001", "02-003", "02-014")),]
count_matrix_8 <- count_matrix_28[, which(coldata_28$SampleID %in% c("01-007", "02-001", "02-003", "02-014"))]


# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

# # Set seed for reproducibility
# set.seed(42)

```

Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix)))` gènes au nom dupliqués, sur un total de `r nrow(count_matrix)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(basic_coldata) == colnames(count_matrix)) == length(rownames(basic_coldata))`.<br/>


# Analyse des 22 samples pre : quels sont les marqueurs de meilleure survie ?

Recherche et analyse des gènes DE des 23 samples pre.<br/>

##  Recherche des DEGs 

Design 1 : ~ Age_cr + Prog_status <br/>
Design 2 : ~ Age_cr + QC_duplication_perc_cr + Prog_status <br/>

```{r DEGs 22 samples}

# coldata_22 <- basic_coldata[basic_coldata$Time == "pre",]
# count_matrix_22 <- count_matrix[, basic_coldata$Time == "pre"]

design_22 <- as.formula("~ Age_cr + Prog_status")
# coef_22 <- "Prog_status_1_vs_0"

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_22,
  colData = coldata_22,
  design = design_22
)

dds <- DESeq(dds)

# norm.count <- as.data.frame(counts(dds_prepost, normalized=TRUE))

# connaître le dernier coef de dds
# resultsNames(dds)
  
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_22 <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

# resLFC <- lfcShrink(dds, coef=coef_23, type="apeglm", lfcThreshold=lfc.cutoff) # apeglm has less bias than "normal"
# /!\ Using contrast forms an expanded model matrix, treating all factor levels equally, and averages over all distances between all pairs of factor levels to estimate the prior. Using coef, means looking only at that column of the model matrix -> à modifier pour utiliser contrast !!
# resLFCordered <- resLFC[order(resLFC$svalue), ]
# selected_genesLFC_22 <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])

# checking the DESeq2 dispersion 
plotDispEsts(dds, main=paste("DESeq2 dispersion plot pre", design_23))
  
# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main=paste("Cooks' distance pre", design_23))
  
# MA-plot
plotMA(resLFC, ylim=c(-2,2), main=paste("plotMA pre", design_22))

# comparaison des designs (à refaire tourner avec 22 samples !!) : avec "~ Age_cr + QC_aligned_M_cr + Prog_status" : 31 DEGs en commun avec le design_22, 197 nouveaux, et 8 seulement avec design_22


```


```{r cross-validation 22 samples}

# Cross-validation to check DEGs reproducibility
cv_samples_DEGs_22 <- function(count_matrix, coldata, design, padj.cutoff=0.05, lfc.cutoff=0.58) { #, coef
  
  # set.seed(123)
  cv_results <- list()
  cv_results_LFC <- list()

  for (outlier in 1:ncol(count_matrix)) {
    col_ind <- 1:ncol(count_matrix)
    indexes <- col_ind[-outlier]
    
    tryCatch({
      # Train on training set
      dds <- DESeqDataSetFromMatrix(
      countData = count_matrix[, indexes],
        colData = coldata[indexes, ],
        design = as.formula(design)
      )
      dds <- DESeq(dds, quiet = TRUE)
      
      res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)
      resOrdered <- res[order(res$padj), ]
      selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
      cv_results[[outlier]] <- selected_genes
      
      # resLFC <- lfcShrink(dds, coef=coef, type="apeglm", lfcThreshold=lfc.cutoff)
      # resLFCordered <- resLFC[order(resLFC$svalue), ]
      # selected_genesLFC <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])
      # cv_results_LFC[[outlier]] <- selected_genesLFC
      
    })
    }
  result <- list('all' = cv_results, 'LFC' = cv_results_LFC)
  return(result)
}

check_repro_22 <- cv_samples_DEGs_22(count_matrix_22, coldata_22, "~ QC_duplication_perc_cr + Age_cr + Best_response_2")#, coef='Prog_status_1_vs_0')

min_nb <- 18
all_DEGs <- do.call(c, check_repro_22$all)
number_by_DEG <- table(all_DEGs)
DEGs_in_common <- list()
for (DEG in rownames(number_by_DEG)) {
  if (number_by_DEG[[DEG]] > min_nb) {
    DEGs_in_common <- append(DEGs_in_common, DEG)
  }
}
DEGs_in_common <- do.call(c, DEGs_in_common)

```



# Analyse des 28 samples : 22 pre vs 6 post

Recherche et analyse des gènes DE entre les 22 samples pre et les 6 samples post.<br/>

Design sans cross-validation : ~ Age_cr + Time <br/>
Design avec cross-validation : ~ QC_duplication_perc_cr + Age_cr + Time <br/>

##  Recherche des DEGs 

```{r DEGs 28 samples}

design_28 <- as.formula("~ Age_cr + Time")
# coef_28 <- "Time_post_vs_pre"

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_28,
  colData = coldata_28,
  design = design_28
)
  
dds$Time <- relevel(dds$Time, ref = "pre")
dds <- DESeq(dds)

# norm.count_28 <- as.data.frame(counts(dds_prepost, normalized=TRUE))
norm_count_28 <- assay(vst(dds, blind=FALSE))
# keep_mean <- rowMeans(norm_counts) > 10
# keep_zero <- rowSums(norm_counts == 0) / ncol(norm_counts) <= 0.90
# dds_filt <- dds[keep_mean & keep_zero, ]
  
res_28 <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)#, contrast=c("Time", "post", "pre")) 
resOrdered_28 <- res_28[order(res_28$padj), ]
selected_genes_28 <- rownames(resOrdered_28[which(resOrdered_28$padj < padj.cutoff), ])

resLFC_28 <- lfcShrink(dds, coef=3, type="apeglm")#, lfcThreshold=lfc.cutoff)#, svalue=FALSE)
# resLFCordered <- resLFC[order(resLFC$svalue), ]
# selected_genesLFC_28 <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])
resLFCordered_28 <- resLFC_28[order(resLFC_28$padj), ]
selected_genesLFC_28 <- rownames(resLFCordered_28[which(resLFCordered_28$padj < padj.cutoff), ])

# checking the DESeq2 dispersion 
plotDispEsts(dds, main=paste("DESeq2 dispersion plot 28", as.character(design_28)[2]))
  
# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main=paste("Cooks' distance 28", as.character(design_28)[2]))

# MA-plot
plotMA(resLFC_28, ylim=c(-2,2), main=paste("plotMA 28", as.character(design_28)[2]))


DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_28, rownames(counts_genes))]
table_28 <- data.frame(selected_genes_28, DEGs_gene_symbols, res_28[match(selected_genes_28, row.names(res_28)), "stat"], res_28[match(selected_genes_28, row.names(res_28)), "padj"], res_28[match(selected_genes_28, row.names(res_28)), "log2FoldChange"])
colnames(table_28) <- c("ENSG", "gene_symbol", "stat", "padj", "log2FoldChange")
table_28$ENSG_clean <- sub("\\..*", "", table_28$ENSG)
table_28 <- table_28[order(table_28$log2FoldChange), ]

# draw DEGs table
as.matrix(table_28) %>%
DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
DT::formatStyle(columns = colnames(table_28),fontSize = '80%')

```


```{r comparaison designs 28}

design_0 <- as.formula("~ Time")
design_1 <- as.formula("~ Age_cr + Time")
design_2 <- as.formula("~ QC_duplication_perc_cr + Age_cr + Time")
design_3 <- as.formula("~ Best_response_2 + QC_duplication_perc_cr + Age_cr + Time")

designs <- list(
  design_0 = design_0,
  design_1 = design_1,
  design_2 = design_2,
  design_3 = design_3
)

selected_genes_list <- list()
norm.count_list <- list()
res_list <- list()
# resLFC_list <- list()
# selected_genesLFC_list_Endo <- list()
for (i in 0:3) {
  design_name <- paste0("design_", i)
  dds <- DESeqDataSetFromMatrix(
    countData = count_matrix_28,
    colData = coldata_28,
    design = designs[[design_name]]
  )
  dds$Time <- relevel(dds$Time, ref = "pre")
  dds <- DESeq(dds)
  # dds_list[[design_name]] <- dds
  
  # norm.count <- as.data.frame(counts(dds_prepost, normalized=TRUE))
  norm.count <- assay(vst(dds, blind=FALSE))
  norm.count_list[[design_name]] <- norm.count
  
  # connaître le dernier coef de dds
  # resultsNames(dds)
  
  res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
  res_list[[design_name]] <- res
  # summary(res_prepost_1)
  resOrdered <- res[order(res$padj), ]
  selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
  selected_genes_list[[design_name]] <- selected_genes
  
}

venn(selected_genes_list)

# Volcano plots without shrink
for ( i in 0:3 ) {
  design_name <- paste0("design_", i)
  with(res_list[[design_name]], plot(log2FoldChange, -log10(padj), pch=20, main=paste("Volcano plot LapNet design", i), cex=0.5))
  with(subset(res_list[[design_name]], padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
  with(subset(res_list[[design_name]], abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
  with(subset(res_list[[design_name]], padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))
}

# heatmaps des différents designs, en colorant par quelques variables d'intérêt
time_col <- c("turquoise","orangered")
names(time_col) <- c('pre', 'post')
# age_col <- rainbow(length(unique(coldata_28$Age_cr)))
# names(age_col) <- sort(unique(coldata_28$Age_cr))
# QC_col <- rainbow(length(unique(coldata_28$QC_duplication_perc_cr)))
# names(QC_col) <- sort(unique(coldata_28$QC_duplication_perc_cr))
# sex_col <- c("pink","cyan")
# names(sex_col) <- c("F","M")
# prog_col <- c("grey","black")
# names(prog_col) <- c("0","1")
# death_col <- c("grey","black")
# names(death_col) <- c("0","1")
# br2_col <- c("grey","black")
# names(br2_col) <- c("PR","SD_PD")

column_ha = HeatmapAnnotation(
  Time=coldata_28$Time,
  # Age=coldata_28$Age_cr, 
  # QC=coldata_28$QC_duplication_perc_cr,
  # Gender=coldata_28$Sex,
  # # Progress=coldata_28$Prog_status,
  # Death=coldata_28$Death_status,
  # Best_response=coldata_28$Best_response_2,
  col=list(Time=time_col#,Gender=sex_col,Best_response=br2_col,Death=death_col#,Age=age_col,QC=QC_col,Progress=prog_col
  ))

distcor <- function(m){ as.dist(1-cor(t(m))  ) }
for ( i in 0:3 ) { 
  design_name <- paste0("design_", i)
  DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_list[[design_name]], rownames(counts_genes))]
  matrix_heatmap <- norm.count_list[[design_name]][match(selected_genes_list[[design_name]], rownames(norm.count_list[[design_name]])),]
  rownames(matrix_heatmap) <- make.names(counts_genes$gene_name[match(rownames(matrix_heatmap), rownames(counts_genes))], unique = T)
  draw(Heatmap(as.matrix(matrix_heatmap),
               clustering_distance_columns  = "pearson",
               clustering_method_columns = "ward.D2",
               clustering_distance_rows  = "pearson",
               clustering_method_rows = "ward.D2",
               top_annotation = column_ha))
  # heatmap.2(as.matrix(matrix_heatmap), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))
}

```


```{r cross-validation 28}

cv_samples_DEGs_28 <- function(count_matrix, coldata, design, padj.cutoff=0.05, lfc.cutoff=0.58) { #, coef
  
  # set.seed(123)
  cv_results <- list()
  cv_results_LFC <- list()

  for (outlier in 1:ncol(count_matrix)) {
    col_ind <- 1:ncol(count_matrix)
    indexes <- col_ind[-outlier]
    
    tryCatch({
      # Train on training set
      dds <- DESeqDataSetFromMatrix(
      countData = count_matrix[, indexes],
        colData = coldata[indexes, ],
        design = as.formula(design)
      )
      dds$Time <- relevel(dds$Time, ref = "pre")
      dds <- DESeq(dds, quiet = TRUE)
      
      res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)
      resOrdered <- res[order(res$padj), ]
      selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
      cv_results[[outlier]] <- selected_genes
      
      # resLFC <- lfcShrink(dds, coef=coef, type="apeglm", lfcThreshold=lfc.cutoff)
      # resLFCordered <- resLFC[order(resLFC$svalue), ]
      # selected_genesLFC <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])
      # cv_results_LFC[[outlier]] <- selected_genesLFC
      
    })
    }
  result <- list('all' = cv_results, 'LFC' = cv_results_LFC)
  return(result)
}

check_repro_28 <- cv_samples_DEGs_28(count_matrix_28, coldata_28, "~ QC_duplication_perc_cr + Age_cr + Time")#, coef='Prog_status_1_vs_0')

min_nb <- 27 # 20
all_DEGs <- do.call(c, check_repro_28$all)
number_by_DEG <- table(all_DEGs)
DEGs_in_common_28 <- list()
for (DEG in rownames(number_by_DEG)) {
  if (number_by_DEG[[DEG]] > min_nb) {
    DEGs_in_common_28 <- append(DEGs_in_common_28, DEG)
  }
}
DEGs_in_common_28 <- do.call(c, DEGs_in_common_28)
DEGs_gene_symbols <- counts_genes$gene_name[match(DEGs_in_common_28, rownames(counts_genes))]

design_28_cv <- as.formula("~ QC_duplication_perc_cr + Age_cr + Time")
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_28,
  colData = coldata_28,
  design = design_28_cv
)
dds$Time <- relevel(dds$Time, ref = "pre")
dds <- DESeq(dds)
norm_count_28 <- assay(vst(dds, blind=FALSE))
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)

matrix_heatmap_28 <- norm_count_28[match(DEGs_in_common_28, rownames(norm_count_28)),]
rownames(matrix_heatmap_28) <- DEGs_gene_symbols
# heatmap.2(as.matrix(matrix_heatmap_28), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))


table_cv_28 <- data.frame(DEGs_in_common_28, DEGs_gene_symbols, res[match(DEGs_in_common_28, row.names(res)), "stat"], res[match(DEGs_in_common_28, row.names(res)), "padj"], res[match(DEGs_in_common_28, row.names(res)), "log2FoldChange"])
colnames(table_cv_28) <- c("ENSG", "gene_symbol", "stat", "padj", "log2FoldChange")
table_cv_28$ENSG_clean <- sub("\\..*", "", table_cv_28$ENSG)
table_cv_28 <- table_cv_28[order(table_cv_28$log2FoldChange), ]

# draw DEGs table
as.matrix(table_cv_28) %>%
DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
DT::formatStyle(columns = colnames(table_cv_28),fontSize = '80%')

# nombre de DEGs en commun selon le design, en prenant min_nb = 28 : 
# ~ Time :                                    5 DEGs :  "ENSG00000096060.16" "ENSG00000140323.7"  "ENSG00000142303.15" "ENSG00000197496.7"  "ENSG00000284454.2" 
# ~ Age_cr + Time :                           10 DEGs : "ENSG00000096060.16" "ENSG00000140323.7"  "ENSG00000142303.15" "ENSG00000197496.7"  "ENSG00000284454.2"  "ENSG00000206513.2"  "ENSG00000236625.8" "ENSG00000263309.6"  "ENSG00000005001.11" "ENSG00000161381.15"
# ~ QC_duplication_perc_cr + Age_cr + Time :  21 DEGs : "ENSG00000096060.16" "ENSG00000140323.7"  "ENSG00000142303.15" "ENSG00000197496.7"  "ENSG00000284454.2"  "ENSG00000206513.2"  "ENSG00000236625.8" "ENSG00000263309.6" "ENSG00000005001.11" "ENSG00000161381.15" "ENSG00000006210.7"  "ENSG00000027001.11" "ENSG00000087128.10" "ENSG00000114554.12" "ENSG00000118242.17" "ENSG00000124313.19"  "ENSG00000125733.19" "ENSG00000135605.14" "ENSG00000137936.19" "ENSG00000235396.5"   "ENSG00000284713.3"
# ~ Best_response_2 + QC_duplication_perc_cr + Age_cr + Time : 29  DEGs (pas tous en commun) : "ENSG00000096060.16" "ENSG00000140323.7" "ENSG00000142303.15" "ENSG00000197496.7" "ENSG00000284454.2" "ENSG00000206513.2"  "ENSG00000236625.8" "ENSG00000263309.6"   "ENSG00000005001.11" "ENSG00000161381.15" "ENSG00000006210.7"  "ENSG00000027001.11" "ENSG00000087128.10" "ENSG00000114554.12" "ENSG00000118242.17" "ENSG00000124313.19" "ENSG00000125733.19" "ENSG00000135605.14" /0\ "ENSG00000090104.13" "ENSG00000101440.11" "ENSG00000103241.7"  "ENSG00000104518.12" "ENSG00000169231.14" "ENSG00000175121.12" "ENSG00000177143.5"  "ENSG00000181222.19" "ENSG00000188100.9"   "ENSG00000226618.9"  "ENSG00000232804.5"  
# ~ Height_cr + QC_duplication_perc_cr + Age_cr + Time :  28 DEGs (pas tous en commun) : "ENSG00000096060.16" "ENSG00000140323.7" "ENSG00000142303.15" "ENSG00000197496.7"  "ENSG00000284454.2" /0\ "ENSG00000005001.11" "ENSG00000161381.15" "ENSG00000006210.7"  "ENSG00000027001.11" "ENSG00000087128.10" "ENSG00000114554.12" "ENSG00000118242.17" "ENSG00000124313.19" "ENSG00000137936.19"  "ENSG00000284713.3" "ENSG00000095713.15" "ENSG00000101440.11" "ENSG00000104518.12"  "ENSG00000163288.14" "ENSG00000169231.14" "ENSG00000183311.16" "ENSG00000186564.6"   "ENSG00000198807.13" "ENSG00000226288.3"  "ENSG00000228964.10" "ENSG00000263956.9" "ENSG00000284218.2"  "ENSG00000287725.2" 
# ~ Sex + QC_duplication_perc_cr + Age_cr + Time :        28 DEGs (pas tous en commun) : "ENSG00000096060.16" "ENSG00000140323.7"  "ENSG00000142303.15" "ENSG00000197496.7"  "ENSG00000206513.2" "ENSG00000124313.19"  "ENSG00000125733.19" "ENSG00000135605.14" "ENSG00000137936.19" /0\ "ENSG00000284454.2" "ENSG00000005001.11" "ENSG00000006210.7"  "ENSG00000011347.11" "ENSG00000027001.11" "ENSG00000087128.10" "ENSG00000087250.10" "ENSG00000099984.12" "ENSG00000101440.11" "ENSG00000111432.5"  "ENSG00000114554.12" "ENSG00000118242.17" "ENSG00000127129.11" "ENSG00000161381.15"  "ENSG00000236625.8"  "ENSG00000269741.5" "ENSG00000275723.4"  "ENSG00000284713.3"  "ENSG00000288796.1"
# ~ SampleID + Time :  43 DEGs : "ENSG00000096060.16" /0\ "ENSG00000142303.15" "ENSG00000284454.2" "ENSG00000161381.15" "ENSG00000004399.14" "ENSG00000053900.12" "ENSG00000068024.19" "ENSG00000090382.7"  "ENSG00000101104.14" "ENSG00000102854.17" "ENSG00000109911.20" "ENSG00000110799.15" "ENSG00000113721.15" "ENSG00000114019.16" "ENSG00000115421.14" "ENSG00000117983.17" "ENSG00000122515.17" "ENSG00000124107.5"  "ENSG00000126822.19" "ENSG00000130635.18" "ENSG00000143367.18" "ENSG00000143772.12" "ENSG00000148346.13" "ENSG00000149591.18" "ENSG00000151388.11" "ENSG00000155093.20" "ENSG00000162591.18" "ENSG00000163359.18" "ENSG00000164597.16" "ENSG00000165699.17" "ENSG00000167315.19" "ENSG00000168631.15" "ENSG00000175538.12" "ENSG00000175820.4"  "ENSG00000184445.12" "ENSG00000197381.19" "ENSG00000204291.13" "ENSG00000215182.8"  "ENSG00000233024.7"  "ENSG00000263956.9"  "ENSG00000266338.8"  "ENSG00000271383.10"  "ENSG00000291995.1"

```

## Etude des DEGs sans cross-validation : {.tabset}

### Volcano plot

```{r Volcano 28}

# without shrink
with(resOrdered_28, plot(log2FoldChange, -log10(padj), pch=20, main="Volcano plot NOshrink", cex=0.5))
with(subset(resOrdered_28, padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
with(subset(resOrdered_28, abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
with(subset(resOrdered_28, padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))

# shrinked
with(resLFCordered_28, plot(log2FoldChange, -log10(padj), pch=20, main="Volcano plot shrinked", cex=0.5))
with(subset(resLFCordered_28, padj<.05 ), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
with(subset(resLFCordered_28, abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="grey", cex=0.5))
with(subset(resLFCordered_28, padj <.05 & abs(log2FoldChange)>0.58), points(log2FoldChange, -log10(padj), pch=20, col="green", cex=0.5))

```

<br/>
<br/>

### Heatmap des DEGs

```{r Heatmap DEG 28}

# heatmap Nico des DEGs par sample
distcor <- function(m){ as.dist(1-cor(t(m))  ) }
DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_28, rownames(counts_genes))]
matrix_heatmap_28 <- norm_count_28[match(selected_genes_28, rownames(norm_count_28)),]
rownames(matrix_heatmap_28) <- make.names(counts_genes$gene_name[match(rownames(matrix_heatmap_28), rownames(counts_genes))], unique = T)
heatmap.2(as.matrix(matrix_heatmap_28), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))

# heatmap Emilie avec ComplexHeatmap (/!\ essayer méthode Roxane !!!)
## cluster colors 
time_col <- c("turquoise","orangered")
names(time_col) <- c('pre', 'post')
age_col <- rainbow(length(unique(coldata_28$Age_cr)))
names(age_col) <- sort(unique(coldata_28$Age_cr))
QC_col <- rainbow(length(unique(coldata_28$QC_duplication_perc_cr)))
names(QC_col) <- sort(unique(coldata_28$QC_duplication_perc_cr))
sex_col <- c("pink","cyan")
names(sex_col) <- c("F","M")
prog_col <- c("grey","black")
names(prog_col) <- c("0","1")
death_col <- c("grey","black")
names(death_col) <- c("0","1")
br2_col <- c("grey","black")
names(br2_col) <- c("PR","SD_PD")

## cluster legend
column_ha = HeatmapAnnotation(
  Time=coldata_28$Time,
  # Age=coldata_28$Age_cr, 
  # QC=coldata_28$QC_duplication_perc_cr,
  Gender=coldata_28$Sex,
  # Progress=coldata_28$Prog_status,
  Death=coldata_28$Death_status,
  Best_response=coldata_28$Best_response_2,
  col=list(Time=time_col,Gender=sex_col,Best_response=br2_col,Death=death_col#,Age=age_col,QC=QC_col,Progress=prog_col
  ))

## heatmap
scaled_mat <- t(scale(t(as.matrix(matrix_heatmap_28))))
h <-  Heatmap(scaled_mat,#norm_count_28[which(rownames(norm_count_28) %in% selected_genes_28),],
              col=colorRamp2(c(-4,0,4), c("blue", "white", "red")),
              clustering_distance_columns  = "spearman",#pearson
              clustering_method_columns = "ward.D2",
              clustering_distance_rows  = "spearman",#pearson
              clustering_method_rows = "ward.D2",
              # row_km = 2, , column_km = 2,
              row_split = 2, column_split = 2,
              top_annotation = column_ha)
draw(h)

```

<br/>
<br/>

### ORA (Over-Enrichment Analysis)

```{r ORA 28}

# defining H, C2 (and C5 ?)
# gsets_msigdb_h <- msigdbr(species = "human", collection = "H")
# msigdb.genes.setsH <- gsets_msigdb_h %>% dplyr::select(gs_name, gene_symbol, ensembl_gene) %>% as.data.frame()
# msigdb.genes.setsH$gs_name <- as.factor(msigdb.genes.setsH$gs_name)
# Hallmark <- as.list(1)
# for (i in 1:nlevels(msigdb.genes.setsH$gs_name)){
#   Hallmark[[i]] <- cbind(msigdb.genes.setsH[which(msigdb.genes.setsH$gs_name == levels(msigdb.genes.setsH$gs_name)[i]), 3])
# }
# names(Hallmark) <- levels(msigdb.genes.setsH$gs_name)
# 
# gsets_msigdb_c2 <- msigdbr(species = "human", collection = "C2", subcollection = "CP:REACTOME")
# msigdb.genes.setsC2reactome <- gsets_msigdb_c2 %>% dplyr::select(gs_name, gene_symbol, ensembl_gene) %>% as.data.frame()
# msigdb.genes.setsC2reactome$gs_name <- as.factor(msigdb.genes.setsC2reactome$gs_name)
# C2 <- as.list(1)
# for (i in 1:nlevels(msigdb.genes.setsC2reactome$gs_name)){
#   C2[[i]] <- cbind(msigdb.genes.setsC2reactome[which(msigdb.genes.setsC2reactome$gs_name == levels(msigdb.genes.setsC2reactome$gs_name)[i]), 3])
# }
# names(C2) <- levels(msigdb.genes.setsC2reactome$gs_name)

gsets_msigdb_h <- msigdbr(species = "human", collection = "H")
gsets_msigdb_c2 <- msigdbr(species = "human", collection = "C2")
gsets_msigdb_c5 <- msigdbr(species = "human", collection = "C5")
msigdb.genes.setsH <- gsets_msigdb_h %>% dplyr::select(gs_name,ensembl_gene) %>% as.data.frame()
msigdb.genes.setsC2 <- gsets_msigdb_c2 %>% dplyr::select(gs_name,ensembl_gene) %>% as.data.frame()
msigdb.genes.setsC5 <- gsets_msigdb_c5 %>% dplyr::select(gs_name,ensembl_gene) %>% as.data.frame()

# using enricher from clusterProfiler 
clustP_query <- function(genes, genesets, universe = NULL, adj_p_cutoff) { 
  result <- list() 
  if (is.vector(genes) && length(genes) > 0 && !all(genes == '')) 
    { 
    universe.gene.pathway <- universe 
    other.genes <- universe.gene.pathway[!universe.gene.pathway %in% genesets$gene_symbol] 
    other.genes <- na.omit(unique(other.genes)) 
    other.genes.df <- cbind(rep("other.genes", length(other.genes)), other.genes) 
    colnames(other.genes.df) <- colnames(genesets) 
    genes.sets.all <- rbind(genesets, other.genes.df)

    res <-
      enricher(
        genes,
        universe = universe,
        TERM2GENE = genes.sets.all,
        qvalueCutoff = adj_p_cutoff,
        maxGSSize=1000
      )
    result <- as.data.frame(res)
  } else {
    warning(paste0('genes must be a non-empty vector of gene names'))
  }
  return(result)
}

# genesets <- data.frame(table_28$ENSG_clean, rep('28_WOcv', length(table_28$ENSG_clean)))
universe <- sub("\\..*", "", rownames(count_matrix_28)) # /!\ à changer : "Plutôt mettre les gènes présents dans la matrice que les coding." (doc PGTeque 23/01/25)
DEGs_up <- table_28[which(table_28$log2FoldChange >0),'ENSG_clean']
DEGs_down <- table_28[which(table_28$log2FoldChange <0),'ENSG_clean']
for (geneset in c(msigdb.genes.setsH, msigdb.genes.setsC2, msigdb.genes.setsC5)) {
  res_Enricher_up <- clustP_query(DEGs_up, geneset, universe, padj.cutoff)
  if (nrow(res_Enricher_up) > 0) {
    
  }
  res_Enricher_down <- clustP_query(DEGs_down, geneset, universe, padj.cutoff)
  if (nrow(res_Enricher_down) > 0) {
    
  }
}
# res <- enricher(genes, universe=universe, TERM2GENE=genes.sets.all,  qvalueCutoff=adj_p_cutoff,maxGSSize=1000)
toto <- enricher(
     sub("\\..*", "", rownames(count_matrix_28)[1:1000]),
     universe = universe,
     TERM2GENE = genes.sets.all,
     qvalueCutoff = padj.cutoff,
     maxGSSize=1000
 )
df <- na.omit(toto[1:10,c('Description', 'GeneRatio', 'RichFactor', 'p.adjust')])
df$p.adjust.num <- -log(as.numeric(df$p.adjust),10)
df <- df[order(df$p.adjust.num),]



# DEGs GO enrichment
# sig_genes <- as.character(pathway[which(pathway$svalue < 0.05),'ENSG_clean'])
sig_genes <- as.character(pathway[which(pathway$padj < 0.05),'ENSG_clean'])
all_genes <- as.character(pathway$ENSG_clean)
ego <- enrichGO(gene = sig_genes, 
                universe = all_genes,
                keyType = "ENSEMBL", # si ça marche pas, taper : keytype
                OrgDb = org.Hs.eg.db, 
                ont = "ALL", # "BP"
                pAdjustMethod = "BH", 
                qvalueCutoff = 0.05, 
                readable = TRUE)
dotplot(ego, showCategory=50)
emapplot(ego, showCategory = 50)

# using fgsea
genes_fg <- sub("\\..*", "", selected_genes_28)
universe_bg <- sub("\\..*", "", rownames(count_matrix_28))
foraRes <- fora(genes=genes_fg, universe=universe_bg, pathways=Hallmark)
topPathways <- foraRes[head(order(pval), n=10), pathway]
plotGseaTable(Hallmark[topPathways], genes_fg, foraRes, 
              gseaParam=0.5)

```

<br/>
<br/>

### GSEA

```{r GSEA 28}

# /!\ for functional analysis tools such as GSEA which require fold change values as input you want to provide shrunken values (https://ciri.gitbiopages.ens-lyon.fr/bibs/carine/rnaseq_analysis_training/Session_6.html)

# DEGs > 0

# à modifier pour séparer en UP et DOWN !!!
pathway <- res_28[!is.na(res_28$padj), ]
pathway$ENSG <- rownames(pathway)
pathway$ENSG_clean <- sub("\\..*", "", pathway$ENSG)
pathway$padjustlog10 <- -log10(pathway$padj)
pathway$rank <- pathway$padjustlog10 * sign(pathway$log2FoldChange) # from https://www.biostars.org/p/9526168/
# pathway <- pathway[order(pathway$log2FoldChange, decreasing = T),]
# rankgene <- deframe(pathway[, c("ENSG_clean", "rank")])
pathway <- pathway[order(pathway$stat, decreasing = T),]
rankgene_1 <- setNames(pathway$stat, pathway$ENSG_clean)
pathway <- pathway[order(pathway$rank, decreasing = T),]
rankgene_2 <- setNames(pathway$rank, pathway$ENSG_clean)
# rankgene <- deframe(pathway[, c("ENSG_clean", "log2FoldChange")])

fgseaRes <- fgsea(pathways = Hallmark, 
                  stats    = rankgene_1,
                  minSize  = 10,#15
                  maxSize  = 1000)#500
# fgseaResTidy <- fgseaRes %>%
#   as_tibble() %>%
#   arrange(desc(NES))
# ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
#   geom_col(aes(fill=padj<0.05)) +
#   coord_flip() +
#   labs(x="Pathway", y="Normalized Enrichment Score",
#        title="Hallmark pathways NES from GSEA") + 
#   theme_minimal()
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(Hallmark[topPathways], rankgene_1, fgseaRes, 
              gseaParam=0.5)
collapsedPathways <- collapsePathways(fgseaRes[order(pval)][padj < 0.01], 
                                      Hallmark, rankgene_1)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway]
plotGseaTable(Hallmark[mainPathways], rankgene_1, fgseaRes, 
              gseaParam = 0.5)

fgseaRes <- fgsea(pathways = C2, 
                  stats    = rankgene_1,
                  minSize  = 10,#15
                  maxSize  = 1000)#500
# fgseaResTidy <- fgseaRes %>%
#   as_tibble() %>%
#   arrange(desc(NES))
# ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
#   geom_col(aes(fill=NES < 0)) +
#   coord_flip() +
#   labs(x="Pathway", y="Normalized Enrichment Score",
#        title="Reactome pathways NES from GSEA") + 
#   theme_minimal()
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(C2[topPathways], rankgene_1, fgseaRes, 
              gseaParam=0.5)
collapsedPathways <- collapsePathways(fgseaRes[order(pval)][padj < 0.01], 
                                      C2, rankgene_1)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway]
plotGseaTable(C2[mainPathways], rankgene_1, fgseaRes, 
              gseaParam = 0.5)

# GO avec un Ridgeplot
gse <- gseGO(geneList=rankgene_1, 
              ont ="ALL", #"BP"
              keyType = "ENSEMBL", 
              nPerm = 10000, 
              minGSSize = 3, 
              maxGSSize = 800, 
              pvalueCutoff = 0.05, 
              verbose = TRUE, 
              OrgDb = org.Hs.eg.db, 
              pAdjustMethod = "BH" # "none"
             )
ridgeplot(gse) + labs(x = "enrichment distribution")
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
# gs <- geneset::getMsigdb(org = "human",category = "H")
# gse <- genGSEA(genelist = rankgene, geneset = gs)



```

<br/>
<br/>

```{r XXX}



```

<br/>
<br/>

## Etude des DEGs avec cross-validation : {.tabset}

XXXX


# Analyse des 4 paires pre - post

Recherche et analyse des gènes DE des 4 paires de samples pre - post.<br/>

Design : ~ Age_cr + QC_aligned_M_cr + Time <br/>

## Recherche des DEGs

```{r DEGs 8 samples}

coldata_8 <- basic_coldata[which(basic_coldata$SampleID %in% c("01-007", "02-001", "02-003", "02-014")),]
count_matrix_8 <- count_matrix[, which(basic_coldata$SampleID %in% c("01-007", "02-001", "02-003", "02-014"))]

design_8 <- as.formula("~ SampleID + Time")
# coef_8 <- "Time_post_vs_pre"

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_8,
  colData = coldata_8,
  design = design_8
)

dds$Time <- relevel(dds$Time, ref = "pre")
dds <- DESeq(dds)

norm.count_8 <- as.data.frame(counts(dds, normalized=TRUE))

res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_8 <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

# resLFC <- lfcShrink(dds, coef=coef_8, type="apeglm")#, lfcThreshold=lfc.cutoff)
# resLFCordered <- resLFC[order(resLFC$padj), ]
# selected_genesLFC_8 <- rownames(resLFCordered[which(resLFCordered$padj < padj.cutoff), ])

# checking the DESeq2 dispersion 
plotDispEsts(dds, main=paste("DESeq2 dispersion plot pairs", design_name))

# looking at the number of outliers
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2, main=paste("Cooks' distance pairs", design_name))

# MA-plot
plotMA(resLFC, ylim=c(-2,2), main=paste("plotMA pairs", design_name))

# venn(selected_genes_list_8)#, main="Overlap of DEGs across designs")
# venn(selected_genesLFC_list_8)


```


```{r cross-validation 8}

cv_samples_DEGs_pairs <- function(count_matrix, coldata, design, padj.cutoff=0.05, lfc.cutoff=0.58) { #, coef
  
  # set.seed(123)
  cv_results <- list()
  cv_results_LFC <- list()

  for (outlier in unique(coldata$SampleID)) {
    # col_ind <- 1:ncol(count_matrix)
    # indexes <- col_ind[-outlier]
    
    tryCatch({
      # Train on training set
      dds <- DESeqDataSetFromMatrix(
      countData = count_matrix[, which(coldata$SampleID != outlier)],
        colData = coldata[which(coldata$SampleID != outlier), ],
        design = as.formula(design)
      )
      dds$Time <- relevel(dds$Time, ref = "pre")
      dds <- DESeq(dds, quiet = TRUE)
      
      res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff)
      resOrdered <- res[order(res$padj), ]
      selected_genes <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])
      cv_results[[outlier]] <- selected_genes
      
      # resLFC <- lfcShrink(dds, coef=coef, type="apeglm", lfcThreshold=lfc.cutoff)
      # resLFCordered <- resLFC[order(resLFC$svalue), ]
      # selected_genesLFC <- rownames(resLFCordered[which(resLFCordered$svalue < padj.cutoff/10), ])
      # cv_results_LFC[[outlier]] <- selected_genesLFC
      
    })
    }
  result <- list('all' = cv_results, 'LFC' = cv_results_LFC)
  return(result)
}

check_repro_8 <- cv_samples_DEGs_pairs(count_matrix_8, coldata_8, "~ SampleID + Time")

min_nb <- 1 #2
all_DEGs <- do.call(c, check_repro_8$all)
number_by_DEG <- table(all_DEGs)
DEGs_in_common <- list()
for (DEG in rownames(number_by_DEG)) {
  if (number_by_DEG[[DEG]] > min_nb) {
    DEGs_in_common <- append(DEGs_in_common, DEG)
  }
}
DEGs_in_common <- do.call(c, DEGs_in_common)

```


```{r Analyse DEGs 8}

DEGs_gene_symbols <- counts_genes$gene_name[match(selected_genes_8, rownames(counts_genes))]

# heatmap des DEGs par sample
distcor <- function(m){ as.dist(1-cor(t(m))  ) }
matrix_heatmap_8 <- norm.count_8[match(selected_genes_8, rownames(norm.count_8)),]
# matrix_heatmap_8$gene_symbol <- counts_genes$gene_name[match(rownames(matrix_heatmap_8), rownames(counts_genes))]
# # correction pour le gène dupliqué (à automatiser !)
# matrix_heatmap_8$gene_symbol[75] = "DDR1."
# # matrix_heatmap_8$ENSG <- rownames(matrix_heatmap_8)
# rownames(matrix_heatmap_8) <- matrix_heatmap_8$gene_symbol  
# matrix_heatmap_8 <- subset(matrix_heatmap_8, select = -gene_symbol)
rownames(matrix_heatmap_8) <- make.names(counts_genes$gene_name[match(rownames(matrix_heatmap_8), rownames(counts_genes))], unique = T)

heatmap.2(as.matrix(matrix_heatmap_8), distfun = distcor, hclustfun = function(d)hclust(d,method="ward.D2"), scale="row", trace="none", dendrogram="both",  col = colorRampPalette(rev(brewer.pal(9, "RdBu")))(50))

# GSEA
gsets_msigdb_h <- msigdbr(species = "human", collection = "H")
msigdb.genes.setsH <- gsets_msigdb_h %>% dplyr::select(gs_name, gene_symbol, ensembl_gene) %>% as.data.frame()
msigdb.genes.setsH$gs_name <- as.factor(msigdb.genes.setsH$gs_name)
Hallmark <- as.list(1)
for (i in 1:nlevels(msigdb.genes.setsH$gs_name)){
  Hallmark[[i]] <- cbind(msigdb.genes.setsH[which(msigdb.genes.setsH$gs_name == levels(msigdb.genes.setsH$gs_name)[i]), 3])
}
names(Hallmark) <- levels(msigdb.genes.setsH$gs_name)

gsets_msigdb_c2 <- msigdbr(species = "human", collection = "C2", subcollection = "CP:REACTOME")
msigdb.genes.setsC2reactome <- gsets_msigdb_c2 %>% dplyr::select(gs_name, gene_symbol, ensembl_gene) %>% as.data.frame()
msigdb.genes.setsC2reactome$gs_name <- as.factor(msigdb.genes.setsC2reactome$gs_name)
C2 <- as.list(1)
for (i in 1:nlevels(msigdb.genes.setsC2reactome$gs_name)){
  C2[[i]] <- cbind(msigdb.genes.setsC2reactome[which(msigdb.genes.setsC2reactome$gs_name == levels(msigdb.genes.setsC2reactome$gs_name)[i]), 3])
}
names(C2) <- levels(msigdb.genes.setsC2reactome$gs_name)

pathway <- resLFC[-which(is.na(resLFC$padj)), ]
pathway$ENSG <- rownames(pathway)
pathway$ENSG_clean <- sub("\\..*", "", pathway$ENSG)
pathway$padjustlog10 <- -log10(pathway$padj)
pathway$rank <- pathway$padjustlog10*(pathway$log2FoldChange/abs(pathway$log2FoldChange))
pathway <- pathway[order(pathway$rank, decreasing = T),]
rankgene <- deframe(pathway[, c("ENSG_clean", "log2FoldChange")])

fgseaRes <- fgsea(pathways = Hallmark, 
                  stats    = rankgene,
                  minSize  = 15,
                  maxSize  = 500)
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))
ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA") + 
  theme_minimal()

fgseaRes <- fgsea(pathways = C2, 
                  stats    = rankgene,
                  minSize  = 15,
                  maxSize  = 500)

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

```


# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
