---
title: "LapNet Choix Design - étude des données avant DESeq2"
author: "Eugénie Modolo"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: kable
    highlight: default
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float: no
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

<!--
SPDX-FileCopyrightText: 2026 Eugenie Modolo <eugenie.modolo@lyon.unicancer.fr>

SPDX-License-Identifier: AGPL-3.0-or-later
-->

```{r setup, results = "hide", include=F}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align='center', out.width='100%', cache = F)

path <- "/home/eugenie-modolo/Documents/Lapnet/"
# setwd(path)
knitr::opts_knit$set(root.dir = path)

# WKDIR <- paste0(path,"/DE_analysis")
# setwd(WKDIR)
```

```{r report colors, results = "hide"}
#color html : https://www.w3schools.com/tags/ref_colornames.asp
text.col      <- 'Black'

darker.col    <- '#666699'
darker.col.2  <- '#8c8cb2'
darker.col.3  <- '#ababc7'
darker.col.4  <- '#d1d1e0'
darker.col.5  <- '#f0f0f5'

lighter.col   <- '#c18d0b'
lighter.col.2 <- "#d4af54"
lighter.col.3 <- "#e3cc91"
```

<style> 
toc {position: fixed; overflow:auto}
body {text-align: justify; color: `r text.col`; font-size: 16px}
a:link {color: `r lighter.col.2`}
a:visited {color: `r lighter.col.2`}
h1.title {color: `r darker.col`  ; font-size: 28px}
h1 {color: `r darker.col`  ; font-size: 24px}
h2 {color: `r lighter.col` ; font-size: 20px}
h3 {color: `r darker.col.2` ; font-size: 18px}
h4 {color: `r text.col` ; font-size: 16px}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: `r darker.col.3`}
.datatables{overflow: auto}
td { white-space: nowrap; }
.nav-pills>li>a {
     color: `r darker.col.3`;}
.nav-pills>li>a:hover, .nav-pills>li>a:focus, .nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus{color: black;background-color: `r darker.col.3`;}
</style>

```{r librairies and functions, results = "hide", include=F}
# BiocManager::install("mixOmics")
library("DESeq2")
library("rtracklayer")
library("SummarizedExperiment")
library("pheatmap")
library("RColorBrewer")
library("DEGreport")
library("apeglm")
library("ggrepel")
library("GGally")
library("gridExtra")
library("mixOmics")

library("dplyr")
library("ggplot2")
library("factoextra")
library("reshape2")
library("tidyr")
library("splines")
library('gplots')
library(viridis)
library(conflicted)
# install.packages("estimate", repos="http://R-Forge.R-project.org")

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Objectif

Rechercher le meilleur design pour la recherche de gènes différentiellement exprimés via DESeq2, sur les échantillons LapNet.<br/>
Pour la recherche du meilleur design, on est obligés de travailler sans les échantillons contrôles, dont les données cliniques sont inconnues. On retire également les deux outliers (06-006_pre & 07-003_post).
<br/>
<br/>

```{r creating dataframes}

# colData
coldata <- read.csv('./colData_LapNet.csv', sep=";")
coldata$Best_response_2 <- case_when(
  coldata$Best_response == "SD" ~ "SD_PD",
  coldata$Best_response == "PD" ~ "SD_PD", 
  coldata$Best_response == "PR" ~ "PR",
  TRUE ~ NA_character_  # for any unexpected values
)
coldata$Type <- case_when(
  startsWith(coldata$SampleID, "Ctrl") == TRUE ~ "control",
  startsWith(coldata$SampleID, "Ctrl") == FALSE ~ "lapnet")

# Adding the Splines columns for tumor size progression
# 0. Normalize tumor size at C0
normalize_by_C0 <- function(coldata,
                            size_cols = c("Size_lesion_C0", "Size_lesion_C4", 
                                        "Size_lesion_C8", "Size_lesion_C12"),
                            time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                        "Days_lesion_C8", "Days_lesion_C12")) {
  
  coldata$Size_norm_C0 <- NA
  coldata$Size_norm_C4 <- NA
  coldata$Size_norm_C8 <- NA
  coldata$Size_norm_C12 <- NA
  
  for (i in 1:nrow(coldata)) {
    factor_norm <- 100 / coldata$Size_lesion_C0[i]
    coldata$Size_norm_C0[i] <- 100
    coldata$Size_norm_C4[i] <- coldata$Size_lesion_C4[i] * factor_norm
    coldata$Size_norm_C8[i] <- coldata$Size_lesion_C8[i] * factor_norm
    coldata$Size_norm_C12[i] <- coldata$Size_lesion_C12[i] * factor_norm
  }
  return(coldata)
}

coldata <- normalize_by_C0(coldata)

# 1. Fit natural splines for each patient
calculate_tumor_spline <- function(coldata, 
                                   size_cols = c("Size_norm_C0", "Size_norm_C4", 
                                                 "Size_norm_C8", "Size_norm_C12"),
                                   time_cols = c("Days_lesion_C0", "Days_lesion_C4", 
                                                 "Days_lesion_C8", "Days_lesion_C12"),
                                   df = 2) {  # degrees of freedom for spline
  
  coldata$tumor_spline_slope <- NA
  coldata$tumor_spline_curvature <- NA
  
  for (i in 1:nrow(coldata)) {
    sizes <- as.numeric(coldata[i, size_cols])
    time_points <- as.numeric(coldata[i, time_cols])
    valid_idx <- !is.na(sizes)
    
    if (sum(valid_idx) >= 3) {  # Need at least 3 points for spline
      valid_sizes <- sizes[valid_idx]
      valid_times <- time_points[valid_idx]
      
      tryCatch({
        # Fit spline
        spline_fit <- lm(valid_sizes ~ ns(valid_times, df = min(df, sum(valid_idx) - 1)))
        
        # Extract overall trend (coefficient of first spline basis)
        coldata$tumor_spline_slope[i] <- coef(spline_fit)[2]
        
        # If enough df, get curvature (second derivative approximation)
        if (length(coef(spline_fit)) > 2) {
          coldata$tumor_spline_curvature[i] <- coef(spline_fit)[3]
        }
      }, error = function(e) {
        # Skip if spline fails
      })
    } else if (sum(valid_idx) == 2) {
      # Fall back to linear for 2 points
      valid_sizes <- sizes[valid_idx]
      valid_times <- time_points[valid_idx]
      lm_fit <- lm(valid_sizes ~ valid_times)
      coldata$tumor_spline_slope[i] <- coef(lm_fit)[2]
    }
  }
  
  return(coldata)
}

coldata <- calculate_tumor_spline(coldata, df = 2)

coldata <- subset( coldata, select = - c(Size_lesion_C0, Size_lesion_C4, Size_lesion_C8, Size_lesion_C12, Days_lesion_C0, Days_lesion_C4, Days_lesion_C8, Days_lesion_C12, Best_response, Size_norm_C0))

# scale and center the numeric columns
numeric_columns <- c('Age', 'Height', 'Weight', 'BMI', 'Size_norm_C4', 'Size_norm_C8', 'Size_norm_C12', 'best_lesion_perc', 'OS', 'PFS', 'Surgery', 'QC_duplication_perc', 'QC_alignedM_Salmon', 'QC_uniq_aligned', 'Purity_Estimate', 'Purity_PUREE', 'tumor_spline_slope', 'tumor_spline_curvature')
for (numeric_col in numeric_columns) {
  coldata[paste0(numeric_col, "_cr")] <- scale(coldata[numeric_col])
}
# convert to factors the non-numeric columns
factor_columns <- colnames(coldata) %!in% numeric_columns & !endsWith(colnames(coldata), '_cr')
coldata[factor_columns] <- lapply(coldata[factor_columns], factor)


# countData (gene counts, not transcript's)
counts_genes <- read.table("./results_rnaseq_NFcore/salmon.merged.gene_counts.tsv", 
                     header = TRUE, 
                     row.names = 1, 
                     sep = "\t",
                     check.names = FALSE)
count_columns <- colnames(counts_genes)[colnames(counts_genes) != "gene_name"]
# we need integers for DESeq2
counts_genes[count_columns] <- round(counts_genes[count_columns])
#counts_genes <- as.matrix(counts_genes)

# keeping only the protein-coding genes, as we used a protocol capturing only exons (???)
gtf <- import('/home/eugenie-modolo/Documents/Reference_files/gencode.v49.primary_assembly.annotation.gtf')
gtf_df <- as.data.frame(gtf)
gene_info <- gtf_df[gtf_df$type == "gene", c("gene_id", "gene_type")]
gene_info <- unique(gene_info)
keep_biotype <- c("protein_coding","IG_V_gene","IG_C_gene","IG_J_gene","IG_D_gene","TR_V_gene","TR_J_gene" ,"TR_C_gene","TR_D_gene")
keep_genes <- gene_info[which(gene_info$gene_type %in% keep_biotype), 'gene_id']
count_matrix_pcg <- counts_genes[which(rownames(counts_genes) %in% keep_genes), ]
count_matrix_pcg <- as.matrix(count_matrix_pcg[, count_columns])

coldata <- data.frame(coldata, row.names = count_columns)

coldata <- subset( coldata, select = -c(Age, Height, Weight, BMI, Size_norm_C4, Size_norm_C8, Size_norm_C12, best_lesion_perc, OS, PFS, Surgery, QC_duplication_perc, QC_alignedM_Salmon, QC_uniq_aligned, Purity_Estimate, Purity_PUREE, tumor_spline_slope, tumor_spline_curvature) )

for (col in colnames(coldata)) {
    if (endsWith(col, '_cr')) {
        coldata[[col]] <- as.numeric(as.vector(coldata[[col]]))
    }
}

coldata_lapnet <- coldata[which(coldata$Type == "lapnet"), ]
coldata_lapnet <- subset( coldata_lapnet, select = -c(Type) ) # colonne qui ne prend qu'une valeur
count_matrix_lapnet <- count_matrix_pcg[, which(coldata$Type == "lapnet")]

# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- 0.58

```

Checkpoints: <br/>
- nous obtenons `r sum(duplicated(rownames(count_matrix_pcg)))` gène(s) au nom dupliqué, sur un total de `r nrow(count_matrix_pcg)` protein-coding gènes.<br/>
- vérification que les samples sont les mêmes (même nombre et même ordre) entre coldata et la matrice de comptage : `r sum(rownames(coldata) == colnames(count_matrix_pcg)) == length(rownames(coldata))`.<br/>


# Peut-on conserver les échantillons avec une faible QC (06-006_pre & 07-003_post) ?

On a vu sur les ACP que ces deux échantillons sont des outliers dès l'axe 1. Faisons des vérifications supplémentaires avant de décider de les retirer définitivement de l'étude des DEGs.

## Nombre de gènes à zéro counts dans chaque échantillon

```{r cheks on 06-006}

coldata_figures <- coldata

# checking number of genes at 0 for each sample
coldata_figures$nber_0 <- NA
coldata_figures$perc_0 <- NA
n_genes <- nrow(count_matrix_pcg)
for (id in 1:ncol(count_matrix_pcg)) { 
  coldata_figures$nber_0[id] <- sum(count_matrix_pcg[,id] == 0) 
  coldata_figures$perc_0[id] <- coldata_figures$nber_0[id] / n_genes
  }

coldata_figures$row.names <- rownames(coldata_figures)

ggplot(coldata_figures, aes(x = reorder(row.names, perc_0), y = perc_0, fill = Time)) +
     geom_col() +
     labs(title = "LapNet - % of 0 counts per Sample",
          x = "Sample",
          y = "Percentage of genes at 0 count") +
     # ylim(0, 1) +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ggplot(coldata_figures, aes(x = reorder(row.names, nber_0), y = nber_0, fill = Time)) +
#      geom_col() +
#      labs(title = "LapNet - Nb of 0 counts per Sample",
#           x = "Sample",
#           y = "Number of genes at 0 count") +
#      theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

On peut voir que ces deux échantillons ont un pourcentage bien plus élevé de gènes à expression nulle.

## Effet de 06-006_pre & 07-003_post sur le nombre de DEGs (design : ~ Time)

```{r venn DEGs with and without outliers}

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_lapnet,
  colData = coldata_lapnet,
  design = ~ Time
)
dds <- DESeq(dds)
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_with <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_lapnet[, rownames(coldata_lapnet) != "06-006_pre"],
  colData = coldata_lapnet[rownames(coldata_lapnet) != "06-006_pre", ],
  design = ~ Time
)
dds <- DESeq(dds)
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_without06 <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_lapnet[, rownames(coldata_lapnet) != "07-003_post"],
  colData = coldata_lapnet[rownames(coldata_lapnet) != "07-003_post", ],
  design = ~ Time
)
dds <- DESeq(dds)
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_without07 <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_lapnet[, rownames(coldata_lapnet) != "06-006_pre" & rownames(coldata_lapnet) != "07-003_post"],
  colData = coldata_lapnet[rownames(coldata_lapnet) != "06-006_pre" & rownames(coldata_lapnet) != "07-003_post", ],
  design = ~ Time
)
dds <- DESeq(dds)
res <- results(dds, alpha=padj.cutoff, lfcThreshold=lfc.cutoff) #, contrast=c("colX", "factor1", "factor2")) 
resOrdered <- res[order(res$padj), ]
selected_genes_withoutboth <- rownames(resOrdered[which(resOrdered$padj < padj.cutoff), ])

pour_venn <- list()
pour_venn[['all_samples']] <- selected_genes_with
pour_venn[['wo_06-006']] <- selected_genes_without06
pour_venn[['wo_07-003']] <- selected_genes_without07
pour_venn[['wo_both']] <- selected_genes_withoutboth
venn(pour_venn)

coldata_restricted <- coldata_lapnet[rownames(coldata_lapnet) != "06-006_pre" & rownames(coldata_lapnet) != "07-003_post", ]
count_matrix_restricted <- count_matrix_lapnet[, rownames(coldata_lapnet) != "06-006_pre" & rownames(coldata_lapnet) != "07-003_post"]

```

Conclusion : de nombreux DEGs apparaissent spécifiquement lorsqu'au moins l'un de ces échantillons est présent, il semble donc préférable de les retirer de l'analyse.


# Utilisation du LRT pour choisir le design : pre vs post

```{r forward selection strategy LRT with duplicates pre post}

# Function to check if a design matrix is full rank
check_design_rank <- function(count_matrix, coldata, formula) {
  
  tryCatch({
    # Create model matrix
    mm <- model.matrix(formula, data = coldata)
    # Check rank
    is_full_rank <- qr(mm)$rank == ncol(mm)
    
    return(list(
      is_valid = is_full_rank,
      rank = qr(mm)$rank,
      ncol = ncol(mm)
    ))
  }, error = function(e) {
    return(list(
      is_valid = FALSE,
      error = e$message
    ))
  })
}

test_variable_influence_lrt_sequential_duplicates_withVarInt <- function(count_matrix,
                                                                         coldata,
                                                                         duplicates,
                                                                         variables = NULL,
                                                                         var_interest = NULL,
                                                                         padj_cutoff = 0.05,
                                                                         lfc_cutoff = 0.58) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    variables <- colnames(coldata)
  } else {
    missing_vars <- dplyr::setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  duplicates_list <- do.call(c, duplicates)
  
  # Storage for all results
  all_rounds_results <- list()
  selected_variables <- var_interest
  remaining_variables <- variables
  
  round_num <- 1
  check_zero_deg <- 1
  
  # Continue until all variables are tested
  while (length(remaining_variables) > 0 & check_zero_deg == 1) {

    # cat("\n========================================\n")
    # cat("ROUND", round_num, "\n")
    # cat("========================================\n")
    
    if (round_num == 1) {
      # cat("Testing variables alone (reduced = ~1)\n\n")
      round_results <- list()
      reduced_formula <- "~1"
      full_formula <- as.formula(paste("~", var_interest))
      rank_check <- check_design_rank(count_matrix, coldata, full_formula)
      dds <- DESeqDataSetFromMatrix(
        countData = count_matrix,
        colData = coldata,
        design = full_formula
      )
      # Run LRT
      dds_lrt <- DESeq(dds, test = "LRT", 
                       reduced = as.formula(reduced_formula), 
                       quiet = TRUE)
      res <- results(dds_lrt)
      
      # Count DEGs
      n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
      n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                        na.rm = TRUE)
      
      round_results[[var_interest]] <- data.frame(
        round = round_num,
        tested_variable = var_interest,
        current_design = "none",
        full_model = paste("~", var_interest),
        reduced_model = reduced_formula,
        n_deg_padj_only = n_deg_padj,
        n_deg_padj_and_lfc = n_deg_both,
        stringsAsFactors = FALSE
      )
      round_df <- do.call(rbind, round_results)
      rownames(round_df) <- NULL
      
      # Select best variable (most DEGs with padj + LFC)
      best_variable <- round_df$tested_variable[var_interest]
      best_n_deg <- round_df$n_deg_padj_and_lfc[var_interest]
      all_rounds_results[[paste0("round_", round_num)]] <- round_df
      
      round_num <- 2
      reduced_formula <- paste("~", var_interest)
      
    } else {
      # cat("Current design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
      # cat("Testing addition of remaining variables\n\n")
      reduced_formula <- paste("~", paste(selected_variables, collapse = " + "))
    }
    
    # Storage for this round
    round_results <- list()
    skipped_tests <- list()
    
    # Test each remaining variable
    for (var in remaining_variables) {

      full_formula_str <- paste("~", paste(c(var, selected_variables), collapse = " + "))
      # cat("Testing:", var, "added to current design\n")
      
      # Check design validity
      full_formula <- as.formula(full_formula_str)
      rank_check <- check_design_rank(count_matrix, coldata, full_formula)
      
      if (!rank_check$is_valid) {
        cat("  SKIPPED: Design matrix not full rank\n")
        skipped_tests[[var]] <- "Not full rank"
        next
      }
      
      tryCatch({
        # Create dds
        dds <- DESeqDataSetFromMatrix(
          countData = count_matrix,
          colData = coldata,
          design = full_formula
        )
        
        # Run LRT
        dds_lrt <- DESeq(dds, test = "LRT", 
                         reduced = as.formula(reduced_formula), 
                         quiet = TRUE)
        res <- results(dds_lrt)
        
        # Count DEGs
        n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
        n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                          na.rm = TRUE)
        
        round_results[[var]] <- data.frame(
          round = round_num,
          tested_variable = var,
          current_design = ifelse(round_num == 1, "none", 
                                  paste(selected_variables, collapse = " + ")),
          full_model = full_formula_str,
          reduced_model = reduced_formula,
          n_deg_padj_only = n_deg_padj,
          n_deg_padj_and_lfc = n_deg_both,
          stringsAsFactors = FALSE
        )
        
        # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
        
      }, error = function(e) {
        cat("  ERROR:", e$message, "\n")
        skipped_tests[[var]] <- e$message
      })
    }
    
    # Combine results for this round
    if (length(round_results) == 0) {
      cat("\nNo successful tests in this round. Stopping.\n")
      break
    }
    
    round_df <- do.call(rbind, round_results)
    rownames(round_df) <- NULL
    
    # Select best variable (most DEGs with padj + LFC)
    best_idx <- order(round_df$n_deg_padj_and_lfc, round_df$n_deg_padj_only, 
                      decreasing = TRUE)[1] # which.max(round_df$n_deg_padj_and_lfc)
    best_variable <- round_df$tested_variable[best_idx]
    best_n_deg <- round_df$n_deg_padj_and_lfc[best_idx]
    
    if (sum(round_df$n_deg_padj_and_lfc) == 0 & sum(round_df$n_deg_padj_only) == 0) {
      cat("\nNo variable gave at least 1 DEG with the LRT method. Stopping.\n")
      check_zero_deg <- 0
    }
    
    # cat("\n--- ROUND", round_num, "SUMMARY ---\n")
    # cat("Best variable:", best_variable, "with", best_n_deg, "DEGs\n")
    
    # Add to results
    all_rounds_results[[paste0("round_", round_num)]] <- round_df
    
    # Update for next round
    selected_variables <- c(best_variable, selected_variables)
    if (best_variable %in% duplicates_list) {
      which_sublist <- which(sapply(duplicates, function(x) best_variable %in% x))
      remaining_variables <- dplyr::setdiff(remaining_variables, duplicates[[which_sublist]])
    }
    else {
      remaining_variables <- dplyr::setdiff(remaining_variables, best_variable)
    }
    
    round_num <- round_num + 1
  }
  
  # Combine all rounds
  final_results <- do.call(rbind, all_rounds_results)
  rownames(final_results) <- NULL
  
  # Summary
  # cat("\n========================================\n")
  # cat("FINAL SUMMARY\n")
  # cat("========================================\n")
  # cat("Variables selected in order:\n")
  for (i in seq_along(selected_variables)) {
    var_info <- final_results[final_results$round == i & 
                                final_results$tested_variable == rev(selected_variables)[i], ]
    # cat(i, ".", rev(selected_variables)[i], "- DEGs:",
    #     var_info$n_deg_padj_and_lfc, "\n")
  }
  
  cat("\nFinal design:", paste("~", paste(selected_variables, collapse = " + ")), "\n")
  
  return(list(
    results = final_results,
    selected_order = rev(selected_variables),
    selection_summary = data.frame(
      rank = 1:length(selected_variables),
      variable = rev(selected_variables),
      n_deg = sapply(rev(selected_variables), function(v) {
        final_results[final_results$tested_variable == v, "n_deg_padj_and_lfc"][1]
      })
    )
  ))
}

# Heatmap showing DEGs for each variable at each selection round
plot_sequential_selection_heatmap <- function(lrt_results, criterion = "padj_and_lfc") {
  
  results_df <- lrt_results$results
  
  # Choose which DEG count to use
  if (criterion == "padj_only") {
    results_df$n_deg <- results_df$n_deg_padj_only
    legend_title <- "DEGs (padj only)"
  } else {
    results_df$n_deg <- results_df$n_deg_padj_and_lfc
    legend_title <- "DEGs (padj + LFC)"
  }
  
  # Order variables by selection order
  variable_order <- lrt_results$selected_order
  if (length(variable_order) < length(unique(results_df$tested_variable))) {
    for (var in unique(results_df$tested_variable)) {
      if (var %!in% variable_order) {
        variable_order <- c(variable_order, var)
      }
    }
  }
  results_df$tested_variable <- factor(results_df$tested_variable, 
                                       levels = variable_order)
  
  # Create the plot
  p <- ggplot(results_df, aes(x = factor(round), y = tested_variable, fill = n_deg)) +
    geom_tile(color = "white", size = 1) +
    geom_text(aes(label = n_deg), color = "white", size = 4, fontface = "bold") +
    scale_fill_viridis(option = "plasma", name = legend_title,
                      na.value = "grey90") +
    labs(title = "Sequential Variable Selection: DEGs per Round",
         subtitle = "Variables ordered by selection rank (down = selected first)",
         x = "Selection Round",
         y = "Variable") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11),
          axis.text.y = element_text(size = 10),
          plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, size = 10),
          legend.position = "right",
          panel.grid = element_blank())
  
  return(p)
}



duplicates <- list()
duplicates$QC <- c("QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr")
duplicates$purity <- c("Purity_Estimate_cr", "Purity_PUREE_cr")
duplicates$status <- c("Death_status", "Prog_status", "Surg_status")
duplicates$ospfs <- c("OS_cr", "PFS_cr", "Surgery_cr")
duplicates$body <- c("Height_cr", "Weight_cr", "BMI_cr")
duplicates$sizelesion <- c("Best_response_2", "Size_norm_C4_cr", "Size_norm_C8_cr", "Size_norm_C12_cr", "best_lesion_perc_cr", "tumor_spline_slope_cr", "tumor_spline_curvature_cr")
duplicates$classif <- c("Classif_collisson", "Classif_moffitt", "Classif_bailey")

# Run sequential selection
sequential_results <- test_variable_influence_lrt_sequential_duplicates_withVarInt(
  count_matrix = count_matrix_restricted,
  coldata = coldata_restricted,
  duplicates = duplicates,
  variables = c("Sex", "Death_status", "Prog_status", "Surg_status", "Classif_collisson", "Classif_moffitt", "Classif_bailey", "Best_response_2", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "Size_norm_C4_cr", "best_lesion_perc_cr", "OS_cr", "PFS_cr", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr", "Purity_Estimate_cr", "Purity_PUREE_cr", "tumor_spline_slope_cr"),  # or NULL for all # removed all variables containing NAs
  var_interest = 'Time',
  padj_cutoff = padj.cutoff,
  lfc_cutoff = lfc.cutoff
)

# View selection order
print(sequential_results$selection_summary)

# Visualize
plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")

```



# Utilisation du LRT pour choisir le design : survie pre


## Choix de la variable d'intérêt

```{r forward selection strategy LRT interest variable pre}

test_variable_influence_lrt <- function(count_matrix,
                                        coldata,
                                        variables = NULL,
                                        padj_cutoff = 0.05,
                                        lfc_cutoff = 0.58,
                                        max_combination_size = 3) {
  
  # Get all variables from coldata if not specified
  if (is.null(variables)) {
    # Try to identify likely variables (exclude sample names, IDs, etc.)
    variables <- colnames(coldata)
    # cat("Available variables in coldata:", paste(variables, collapse = ", "), "\n")
    # cat("Testing all variables\n")
  } else {
    # Check that all specified variables exist
    missing_vars <- dplyr::setdiff(variables, colnames(coldata))
    if (length(missing_vars) > 0) {
      stop("These variables not found in coldata: ", paste(missing_vars, collapse = ", "))
    }
  }
  
  # cat("\nVariables to test:", paste(variables, collapse = ", "), "\n")
  # cat("padj cutoff:", padj_cutoff, "\n")
  # cat("LFC cutoff:", lfc_cutoff, "\n\n")
  
  # Storage for results
  results_list <- list()
  skipped_tests <- list()
  
  # 1. Test each variable alone (reduced = ~1)
  # cat("=== Testing individual variables ===\n")
  for (var in variables) {
    # cat("Testing:", var, "vs ~1\n")
    
    # Check design validity
    formula_full <- as.formula(paste("~", var))
    rank_check <- check_design_rank(count_matrix, coldata, formula_full)
    
    if (!rank_check$is_valid) {
      # cat("  SKIPPED: Design matrix not full rank\n")
      skipped_tests[[paste(var, "vs", "~1")]] <- "Not full rank"
      next
    }
    
    tryCatch({
      # Create dds for this test
      dds <- DESeqDataSetFromMatrix(
        countData = count_matrix,
        colData = coldata,
        design = formula_full
      )
      
      # Run LRT
      dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ 1, quiet = TRUE)
      res <- results(dds_lrt)
      
      # Count DEGs
      n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
      n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                       na.rm = TRUE)
      
      results_list[[var]] <- data.frame(
        full_model = var,
        reduced_model = "~1",
        test_type = "single_variable",
        tested_variable = var,
        n_variables = 1,
        n_deg_padj_only = n_deg_padj,
        n_deg_padj_and_lfc = n_deg_both,
        stringsAsFactors = FALSE
      )
      
      # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
      
    }, error = function(e) {
      cat("  ERROR:", e$message, "\n")
      skipped_tests[[paste(var, "vs", "~1")]] <- e$message
    })
  }
  
  # 2. Test pairs of variables (full = A + B, reduced = A or B)
  if (length(variables) >= 2 && max_combination_size >= 2) {
    # cat("\n=== Testing pairs of variables ===\n")
    
    for (i in 1:(length(variables) - 1)) {
      for (j in (i + 1):length(variables)) {
        var1 <- variables[i]
        var2 <- variables[j]
        
        # Test A + B vs A (to see effect of B)
        # cat("Testing:", var1, "+", var2, "vs", var1, "\n")
        
        formula_full <- as.formula(paste("~", var1, "+", var2))
        formula_reduced <- as.formula(paste("~", var1))
        
        # Check design validity
        rank_check_full <- check_design_rank(count_matrix, coldata, formula_full)
        rank_check_reduced <- check_design_rank(count_matrix, coldata, formula_reduced)
        
        if (!rank_check_full$is_valid || !rank_check_reduced$is_valid) {
          # cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                           na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var1, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var1,
            test_type = "added_variable",
            tested_variable = var2,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
          
        }, error = function(e) {
          # cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var1)]] <- e$message
        })
        
        # Test A + B vs B (to see effect of A)
        # cat("Testing:", var1, "+", var2, "vs", var2, "\n")
        
        formula_reduced2 <- as.formula(paste("~", var2))
        rank_check_reduced2 <- check_design_rank(count_matrix, coldata, formula_reduced2)
        
        if (!rank_check_full$is_valid || !rank_check_reduced2$is_valid) {
          # cat("  SKIPPED: Design matrix not full rank\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- "Not full rank"
          next
        }
        
        tryCatch({
          dds <- DESeqDataSetFromMatrix(
            countData = count_matrix,
            colData = coldata,
            design = formula_full
          )
          
          dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced2, quiet = TRUE)
          res <- results(dds_lrt)
          
          n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
          n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                           na.rm = TRUE)
          
          results_list[[paste(var1, var2, "vs", var2, sep = "_")]] <- data.frame(
            full_model = paste(var1, "+", var2),
            reduced_model = var2,
            test_type = "added_variable",
            tested_variable = var1,
            n_variables = 2,
            n_deg_padj_only = n_deg_padj,
            n_deg_padj_and_lfc = n_deg_both,
            stringsAsFactors = FALSE
          )
          
          # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
          
        }, error = function(e) {
          cat("  ERROR:", e$message, "\n")
          skipped_tests[[paste(var1, var2, "vs", var2)]] <- e$message
        })
      }
    }
  }
  
  # 3. Test triplets (full = A + B + C, reduced = A + B)
  if (length(variables) >= 3 && max_combination_size >= 3) {
    # cat("\n=== Testing triplets of variables ===\n")
    
    for (i in 1:(length(variables) - 2)) {
      for (j in (i + 1):(length(variables) - 1)) {
        for (k in (j + 1):length(variables)) {
          var1 <- variables[i]
          var2 <- variables[j]
          var3 <- variables[k]
          
          # cat("Testing:", var1, "+", var2, "+", var3, "vs", var1, "+", var2, "\n")
          
          formula_full <- as.formula(paste("~", var1, "+", var2, "+", var3))
          formula_reduced <- as.formula(paste("~", var1, "+", var2))
          
          # Check design validity
          rank_check_full <- check_design_rank(count_matrix, coldata, formula_full)
          rank_check_reduced <- check_design_rank(count_matrix, coldata, formula_reduced)
          
          if (!rank_check_full$is_valid || !rank_check_reduced$is_valid) {
            # cat("  SKIPPED: Design matrix not full rank\n")
            skipped_tests[[paste(var1, var2, var3, "vs", var1, var2)]] <- "Not full rank"
            next
          }
          
          tryCatch({
            dds <- DESeqDataSetFromMatrix(
              countData = count_matrix,
              colData = coldata,
              design = formula_full
            )
            
            dds_lrt <- DESeq(dds, test = "LRT", reduced = formula_reduced, quiet = TRUE)
            res <- results(dds_lrt)
            
            n_deg_padj <- sum(res$padj < padj_cutoff, na.rm = TRUE)
            n_deg_both <- sum(res$padj < padj_cutoff & abs(res$log2FoldChange) >= lfc_cutoff, 
                             na.rm = TRUE)
            
            results_list[[paste(var1, var2, var3, "vs", var1, var2, sep = "_")]] <- data.frame(
              full_model = paste(var1, "+", var2, "+", var3),
              reduced_model = paste(var1, "+", var2),
              test_type = "added_to_pair",
              tested_variable = var3,
              n_variables = 3,
              n_deg_padj_only = n_deg_padj,
              n_deg_padj_and_lfc = n_deg_both,
              stringsAsFactors = FALSE
            )
            
            # cat("  DEGs (padj):", n_deg_padj, "| DEGs (padj + LFC):", n_deg_both, "\n")
            
          }, error = function(e) {
            cat("  ERROR:", e$message, "\n")
            skipped_tests[[paste(var1, var2, var3, "vs", var1, var2)]] <- e$message
          })
        }
      }
    }
  }
  
  # Combine all results
  if (length(results_list) == 0) {
    cat("\nWARNING: No tests were successful!\n")
    results_df <- data.frame()
  } else {
    results_df <- do.call(rbind, results_list)
    rownames(results_df) <- NULL
  }
  
  # Summary
  # cat("\n=== SUMMARY ===\n")
  # cat("Successful tests:", length(results_list), "\n")
  # cat("Skipped/failed tests:", length(skipped_tests), "\n")
  
  if (length(skipped_tests) > 0) {
    cat("\nSkipped tests:\n")
    for (test_name in names(skipped_tests)) {
      cat("  ", test_name, ":", skipped_tests[[test_name]], "\n")
    }
  }
  
  return(list(
    results = results_df,
    skipped = skipped_tests
  ))
}


# Run sequential selection
lrt_results <- test_variable_influence_lrt(
  count_matrix = count_matrix_restricted,
  coldata = coldata_restricted,
  variables = c("Death_status", "Prog_status", "Surg_status", "Best_response_2", "Size_norm_C4_cr", "best_lesion_perc_cr", "OS_cr", "PFS_cr", "tumor_spline_slope_cr"),  # or NULL for all
  padj_cutoff = padj.cutoff,
  lfc_cutoff = lfc.cutoff,
  max_combination_size = 1  # Test singles only
)

# draw the results table
as.matrix(lrt_results$results) %>%
DT::datatable(rownames=F, escape = F, width ='100%',filter="top") %>%
DT::formatStyle(columns = colnames(lrt_results$results),fontSize = '80%')

```

La meilleure variable d'intérêt semble être tumor_spline_slope. En PLS, elle était également bien placée. 
Elle est cependant éloignée des variables habituellement surveillées, on va donc également tester OS (qui est mieux placée que PFS).

## Choix du reste du design

### Avec tumor_spline_slope

```{r forward selection strategy LRT with duplicates pre tumor_spline_slope}

# with tumor_spline_slope_cr
sequential_results <- test_variable_influence_lrt_sequential_duplicates_withVarInt(
  count_matrix = count_matrix_restricted,
  coldata = coldata_restricted,
  duplicates = duplicates,
  variables = c("Time", "Sex", "Death_status", "Prog_status", "Surg_status", "Classif_collisson", "Classif_moffitt", "Classif_bailey", "Best_response_2", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "Size_norm_C4_cr", "best_lesion_perc_cr", "OS_cr", "PFS_cr", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr", "Purity_Estimate_cr", "Purity_PUREE_cr"),  # or NULL for all
  var_interest = 'tumor_spline_slope_cr',
  padj_cutoff = padj.cutoff,
  lfc_cutoff = lfc.cutoff
)
plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")

```

### Avec OS

```{r forward selection strategy LRT with duplicates pre OS}

sequential_results <- test_variable_influence_lrt_sequential_duplicates_withVarInt(
  count_matrix = count_matrix_restricted,
  coldata = coldata_restricted,
  duplicates = duplicates,
  variables = c("Time", "Sex", "Death_status", "Prog_status", "Surg_status", "Classif_collisson", "Classif_moffitt", "Classif_bailey", "Best_response_2", "Age_cr", "Height_cr", "Weight_cr", "BMI_cr", "Size_norm_C4_cr", "best_lesion_perc_cr", "PFS_cr", "QC_duplication_perc_cr", "QC_alignedM_Salmon_cr", "QC_uniq_aligned_cr", "Purity_Estimate_cr", "Purity_PUREE_cr", "tumor_spline_slope_cr"),  # or NULL for all
  var_interest = 'OS_cr',
  padj_cutoff = padj.cutoff,
  lfc_cutoff = lfc.cutoff
)
plot_sequential_selection_heatmap(sequential_results, criterion = "padj_and_lfc")

```




# Conclusion et remarques

Le choix définitif du design sera fait dans l'étude des DEGs. <br/>



# Versions of all tools used in this analysis

```{r session info}

sessionInfo()

```
